---
title: "0.3 QC and Baseline SuSiE Screen"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
editor_options:
  markdown:
    wrap: 72
---

This workbook performs a hard-QC pass over the Phase 1 GTEx Lung locus screen,
then writes a Slurm task manifest for vanilla `susieR::susie_rss()` baseline
screening. It does not use AlphaGenome annotations and does not run the full
baseline screen inside the workbook.

```{r setup, message=FALSE}
required_packages <- c("arrow", "dplyr", "ggplot2", "readr", "tibble", "tidyr", "purrr")
missing_packages <- required_packages[!vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing_packages) > 0) {
  stop(
    "Missing required R packages: ",
    paste(missing_packages, collapse = ", "),
    ". Install them before running this workbook."
  )
}

library(arrow)
library(dplyr)
library(ggplot2)
library(readr)
library(tibble)
library(tidyr)
library(purrr)
```

# 1. Configuration

```{r config}
find_project_root <- function(start = getwd()) {
  # Example fallback path -- set to your environment (or set EQTL_SUSINE_PROJECT_ROOT).
  hardcoded <- "/path/to/eQTL_annotations_for_susine"
  env_root <- Sys.getenv("EQTL_SUSINE_PROJECT_ROOT", unset = "")
  candidates <- unique(c(
    env_root,
    hardcoded,
    normalizePath(start, winslash = "/", mustWork = FALSE),
    dirname(normalizePath(start, winslash = "/", mustWork = FALSE)),
    dirname(dirname(normalizePath(start, winslash = "/", mustWork = FALSE)))
  ))
  candidates <- candidates[nzchar(candidates)]
  for (candidate in candidates) {
    if (
      dir.exists(file.path(candidate, "utils")) &&
      dir.exists(file.path(candidate, "vignettes")) &&
      file.exists(file.path(candidate, "README.md"))
    ) {
      return(normalizePath(candidate, winslash = "/", mustWork = TRUE))
    }
  }
  stop("Could not locate eQTL_annotations_for_susine project root.")
}

PROJECT_ROOT <- find_project_root()

MANIFEST_PATH <- file.path(PROJECT_ROOT, "config", "loci_manifest_sample_100_per_chrom.csv")
PHASE1_METRICS_CANDIDATES <- c(
  file.path(PROJECT_ROOT, "output", "prelim", "loci_manifest_sample_100_per_chrom_phase1_dataset_metrics.csv"),
  file.path(PROJECT_ROOT, "output", "prelim", "phase1_dataset_metrics_all_loci.csv")
)
PHASE1_METRICS_PATH <- PHASE1_METRICS_CANDIDATES[file.exists(PHASE1_METRICS_CANDIDATES)][1]
if (is.na(PHASE1_METRICS_PATH)) {
  stop(
    "Could not find a Phase 1 metrics CSV. Checked:\n",
    paste(PHASE1_METRICS_CANDIDATES, collapse = "\n")
  )
}

OUTPUT_DIR <- file.path(PROJECT_ROOT, "output")
LD_DIR <- file.path(OUTPUT_DIR, "ld")
SCREEN_DIR <- file.path(OUTPUT_DIR, "prelim", "baseline_susie_screen")
TASK_OUTPUT_ROOT <- file.path(OUTPUT_DIR, "baseline_susie_screen", "task_outputs")
SLURM_SCRIPT_DIR <- file.path(OUTPUT_DIR, "slurm_scripts")
SLURM_PRINTS_DIR <- file.path(OUTPUT_DIR, "slurm_prints")
RUN_TASK_SCRIPT <- file.path(PROJECT_ROOT, "scripts", "run_baseline_susie_screen_task.R")
RUN_PHASE1_TASK_SCRIPT <- file.path(PROJECT_ROOT, "scripts", "run_phase1_task.py")

dir.create(SCREEN_DIR, recursive = TRUE, showWarnings = FALSE)
dir.create(TASK_OUTPUT_ROOT, recursive = TRUE, showWarnings = FALSE)
dir.create(SLURM_SCRIPT_DIR, recursive = TRUE, showWarnings = FALSE)
dir.create(SLURM_PRINTS_DIR, recursive = TRUE, showWarnings = FALSE)

QC_RECONSTRUCT_LD <- TRUE
MAX_VARIANTS_FOR_SUSIE <- Inf
TASK_LIMIT <- Inf
ALLOW_BASELINE_WITH_MISSING_PHASE1_ARTIFACTS <- FALSE

SUSIE_L <- 10L
ACTIVE_ALPHA_MAX_THRESHOLD <- 0.01

JOB_NAME <- "baseline_susie_screen"
TIME_LIMIT <- "03:00:00"
MEMORY <- "10G"
CPUS_PER_TASK <- 1L
ACCOUNT <- "your_slurm_account"  # example -- set to your environment
MAIL_USER <- "you@example.com"  # example -- set to your environment
MAIL_TYPE <- "BEGIN,END,FAIL"
LOAD_R_MODULE <- TRUE
R_LIBS_USER <- "~/R/x86_64-pc-linux-gnu-library/4.3"  # example -- set to your environment
PYTHON_BIN <- "python"

QC_TABLE_PATH <- file.path(SCREEN_DIR, "qc_locus_table.csv")
TASK_MANIFEST_PATH <- file.path(SCREEN_DIR, "baseline_susie_task_manifest.csv")
SLURM_SCRIPT_PATH <- file.path(SLURM_SCRIPT_DIR, "baseline_susie_screen.slurm")
PHASE1_MATERIALIZATION_MANIFEST_PATH <- file.path(SCREEN_DIR, "baseline_susie_missing_phase1_artifacts_manifest.csv")
PHASE1_MATERIALIZATION_SLURM_PATH <- file.path(SLURM_SCRIPT_DIR, "baseline_susie_materialize_phase1_artifacts.slurm")

cat("Project root:", PROJECT_ROOT, "\n")
cat("Manifest:", MANIFEST_PATH, "\n")
cat("Phase 1 metrics:", PHASE1_METRICS_PATH, "\n")
cat("Screen dir:", SCREEN_DIR, "\n")
cat("Task script:", RUN_TASK_SCRIPT, "\n")
```

# 2. QC Helpers

```{r helpers}
accept_phase1_status <- function(status) {
  status %in% c("completed", "skipped_existing")
}

required_input_paths <- function(locus_id, gene_name) {
  locus_dir <- file.path(LD_DIR, locus_id)
  tibble(
    master_path = file.path(locus_dir, sprintf("%s_phase1_master_variants.csv", gene_name)),
    order_path = file.path(locus_dir, sprintf("%s_LD_variant_order.tsv", gene_name)),
    ld_long_path = file.path(locus_dir, sprintf("%s_phase1_LD_R_long.parquet", gene_name))
  )
}

coalesce_path <- function(primary, fallback) {
  primary <- as.character(primary)
  if (length(primary) == 0 || is.na(primary[[1]]) || !nzchar(primary[[1]])) fallback else primary[[1]]
}

row_value <- function(row, name) {
  if (name %in% names(row)) row[[name]] else NA_character_
}

phase1_input_paths <- function(row) {
  fallback <- required_input_paths(row$locus_id[[1]], row$gene_name[[1]])
  tibble(
    master_path = coalesce_path(row_value(row, "phase1_master_variants_csv"), fallback$master_path[[1]]),
    order_path = coalesce_path(row_value(row, "ld_variant_order_tsv"), fallback$order_path[[1]]),
    ld_long_path = coalesce_path(row_value(row, "phase1_ld_long_parquet"), fallback$ld_long_path[[1]])
  )
}

append_issue <- function(issues, condition, label) {
  if (isTRUE(condition)) c(issues, label) else issues
}

qc_one_locus <- function(row) {
  locus_id <- row$locus_id[[1]]
  gene_name <- row$gene_name[[1]]
  phase1_status <- if ("status" %in% names(row)) row$status[[1]] else NA_character_
  n_variants_raw <- if ("phase1_master_variants" %in% names(row)) {
    row$phase1_master_variants[[1]]
  } else {
    NA_integer_
  }
  paths <- phase1_input_paths(row)
  issues <- character()

  issues <- append_issue(issues, !accept_phase1_status(phase1_status), "bad_phase1_status")
  issues <- append_issue(issues, !file.exists(paths$master_path[[1]]), "missing_master")
  issues <- append_issue(issues, !file.exists(paths$order_path[[1]]), "missing_order")
  issues <- append_issue(issues, !file.exists(paths$ld_long_path[[1]]), "missing_ld_long")

  n_variants <- suppressWarnings(as.integer(n_variants_raw))
  issues <- append_issue(issues, is.na(n_variants) || n_variants < 1L, "no_phase1_master_variants")
  issues <- append_issue(
    issues,
    is.finite(MAX_VARIANTS_FOR_SUSIE) && !is.na(n_variants) && n_variants > MAX_VARIANTS_FOR_SUSIE,
    "too_large_for_baseline_susie"
  )

  n_rss <- NA_integer_
  deep_error <- NA_character_

  if (length(issues) == 0 && isTRUE(QC_RECONSTRUCT_LD)) {
    tryCatch({
      master_df <- readr::read_csv(paths$master_path[[1]], show_col_types = FALSE)
      if ("ld_included" %in% names(master_df)) {
        master_df <- master_df %>% dplyr::filter(.data$ld_included)
      }
      master_df <- master_df %>%
        dplyr::transmute(
          variant_id = as.character(.data$variant_id),
          z_score = as.numeric(.data$z_score),
          sample_size = as.numeric(.data$sample_size)
        )

      order_df <- readr::read_tsv(paths$order_path[[1]], show_col_types = FALSE) %>%
        dplyr::mutate(index = as.integer(.data$index), id = as.character(.data$id)) %>%
        dplyr::arrange(.data$index)

      aligned_df <- order_df %>%
        dplyr::left_join(master_df, by = c("id" = "variant_id")) %>%
        dplyr::arrange(.data$index)
      p <- nrow(aligned_df)

      issues <<- append_issue(issues, p < 1L, "no_aligned_variants")
      issues <<- append_issue(issues, !all(aligned_df$index == seq_len(p) - 1L), "noncontiguous_ld_order")
      issues <<- append_issue(issues, any(is.na(aligned_df$z_score)), "missing_aligned_z")
      issues <<- append_issue(issues, any(!is.finite(aligned_df$z_score)), "nonfinite_aligned_z")
      issues <<- append_issue(
        issues,
        any(is.na(aligned_df$sample_size)) || any(!is.finite(aligned_df$sample_size)),
        "invalid_sample_size"
      )
      issues <<- append_issue(issues, any(aligned_df$sample_size <= 2, na.rm = TRUE), "sample_size_le_2")

      ld_long_df <- arrow::read_parquet(paths$ld_long_path[[1]]) %>% tibble::as_tibble()
      expected_pairs <- p * (p - 1) / 2
      issues <<- append_issue(issues, nrow(ld_long_df) != expected_pairs, "ld_pair_count_mismatch")
      issues <<- append_issue(
        issues,
        any(!is.finite(as.numeric(ld_long_df$r))),
        "nonfinite_ld_r"
      )
      issues <<- append_issue(
        issues,
        any(ld_long_df$snp_index_1 < 0 | ld_long_df$snp_index_1 >= p |
              ld_long_df$snp_index_2 < 0 | ld_long_df$snp_index_2 >= p, na.rm = TRUE),
        "ld_index_out_of_range"
      )

      n_rss <<- round(mean(aligned_df$sample_size, na.rm = TRUE))
    }, error = function(e) {
      issues <<- c(issues, "deep_qc_error")
      deep_error <<- conditionMessage(e)
    })
  }

  qc_status <- if (length(issues) == 0) "qc_pass" else "qc_fail"
  dplyr::bind_cols(
    row,
    paths,
    tibble(
      n_variants = n_variants,
      n_rss_qc = n_rss,
      qc_status = qc_status,
      qc_issue_count = length(issues),
      qc_issues = if (length(issues)) paste(issues, collapse = ";") else NA_character_,
      qc_error_message = deep_error
    )
  )
}
```

# 3. Load Phase 1 Screen

```{r load-phase1}
manifest_df <- readr::read_csv(MANIFEST_PATH, show_col_types = FALSE)
phase1_df <- readr::read_csv(PHASE1_METRICS_PATH, show_col_types = FALSE)
manifest_required_cols <- c(
  "locus_id", "gene_name", "gene_id", "gene_type",
  "gtex_tissue", "gtex_chrom", "enabled"
)
missing_manifest_cols <- setdiff(manifest_required_cols, names(manifest_df))
if (length(missing_manifest_cols)) {
  stop("Manifest is missing required columns: ", paste(missing_manifest_cols, collapse = ", "))
}

screen_source_df <- manifest_df %>%
  dplyr::left_join(
    phase1_df,
    by = c("locus_id", "gene_name", "gene_id", "gtex_tissue", "gtex_chrom")
  ) %>%
  dplyr::filter(isTRUE(.data$enabled) | .data$enabled %in% c(TRUE, "True", "TRUE", 1L))

cat("Enabled manifest loci:", nrow(screen_source_df), "\n")
cat("Phase 1 rows:", nrow(phase1_df), "\n")
```

# 4. Run Hard QC

```{r qc-pass}
qc_locus_table <- purrr::map_dfr(seq_len(nrow(screen_source_df)), function(i) {
  if (i %% 25L == 0L || i == nrow(screen_source_df)) {
    cat("QC locus", i, "of", nrow(screen_source_df), "\n")
  }
  qc_one_locus(screen_source_df[i, , drop = FALSE])
})

readr::write_csv(qc_locus_table, QC_TABLE_PATH)

qc_locus_table %>%
  dplyr::count(.data$qc_status, .data$qc_issues, sort = TRUE)
```

```{r qc-summary}
stopifnot(nrow(qc_locus_table) == nrow(screen_source_df))
stopifnot(!anyDuplicated(qc_locus_table$locus_id))

qc_locus_table %>%
  dplyr::count(.data$qc_status)

artifact_missing_loci <- qc_locus_table %>%
  dplyr::mutate(
    missing_master_file = !file.exists(.data$master_path),
    missing_order_file = !file.exists(.data$order_path),
    missing_ld_long_file = !file.exists(.data$ld_long_path),
    missing_any_phase1_artifact = .data$missing_master_file | .data$missing_order_file | .data$missing_ld_long_file,
    recoverable_missing_phase1_artifacts =
      accept_phase1_status(.data$status) &
      !is.na(.data$n_variants) &
      .data$n_variants > 0L &
      .data$missing_any_phase1_artifact
  ) %>%
  dplyr::filter(.data$recoverable_missing_phase1_artifacts)

if (nrow(artifact_missing_loci) > 0) {
  materialization_cols <- intersect(names(manifest_df), names(artifact_missing_loci))
  phase1_materialization_manifest <- artifact_missing_loci %>%
    dplyr::select(dplyr::all_of(materialization_cols)) %>%
    dplyr::mutate(enabled = TRUE)
  readr::write_csv(phase1_materialization_manifest, PHASE1_MATERIALIZATION_MANIFEST_PATH)

  phase1_slurm_text <- sprintf(
'#!/bin/bash
#SBATCH --job-name=baseline_susie_phase1_materialize
#SBATCH --array=1-%d
#SBATCH --time=%s
#SBATCH --mem=%s
#SBATCH --cpus-per-task=%d
#SBATCH --mail-user=%s
#SBATCH --mail-type=%s
#SBATCH --output=/dev/null
#SBATCH --error=/dev/null
#SBATCH --account=%s

set -euo pipefail

PROJECT_ROOT="%s"
MANIFEST_PATH="%s"
RUN_TASK_SCRIPT="%s"
SLURM_PRINTS_BASE="%s"

JOBNAME="${SLURM_JOB_NAME}"
PARENT_ID="${SLURM_ARRAY_JOB_ID:-$SLURM_JOB_ID}"
TASK_ID="${SLURM_ARRAY_TASK_ID:-1}"
PRINTS_DIR="${SLURM_PRINTS_BASE}/${JOBNAME}/${PARENT_ID}"
mkdir -p "${PRINTS_DIR}"

exec >"${PRINTS_DIR}/${TASK_ID}.out" 2>"${PRINTS_DIR}/${TASK_ID}.err"

echo "[$(date -Is)] Materializing Phase-1 artifacts for task ${TASK_ID}"

%s "${RUN_TASK_SCRIPT}" \\
  --manifest "${MANIFEST_PATH}" \\
  --task-id "${TASK_ID}" \\
  --chunk-size 1 \\
  --run-mode processing \\
  --force

echo "[$(date -Is)] Completed Phase-1 materialization task ${TASK_ID}"
',
    nrow(phase1_materialization_manifest),
    TIME_LIMIT,
    MEMORY,
    CPUS_PER_TASK,
    MAIL_USER,
    MAIL_TYPE,
    ACCOUNT,
    PROJECT_ROOT,
    PHASE1_MATERIALIZATION_MANIFEST_PATH,
    RUN_PHASE1_TASK_SCRIPT,
    SLURM_PRINTS_DIR,
    PYTHON_BIN
  )
  writeLines(phase1_slurm_text, PHASE1_MATERIALIZATION_SLURM_PATH)

  cat("Recoverable loci missing Phase-1 z/order/LD artifacts:", nrow(artifact_missing_loci), "\n")
  cat("Wrote Phase-1 materialization manifest:", PHASE1_MATERIALIZATION_MANIFEST_PATH, "\n")
  cat("Wrote Phase-1 materialization Slurm script:", PHASE1_MATERIALIZATION_SLURM_PATH, "\n")
  cat("Submit first with:\n")
  cat("`sbatch ", PHASE1_MATERIALIZATION_SLURM_PATH, "`\n", sep = "")
}

if (
  nrow(artifact_missing_loci) > 0 &&
  !isTRUE(ALLOW_BASELINE_WITH_MISSING_PHASE1_ARTIFACTS)
) {
  stop(
    "Baseline SuSiE screening needs per-locus Phase-1 artifacts, but ",
    nrow(artifact_missing_loci),
    " recoverable loci are missing master/order/LD files. ",
    "Submit the materialization Slurm script above, then rerun this workbook."
  )
}
```

# 5. Write Baseline SuSiE Task Manifest

```{r task-manifest}
task_manifest <- qc_locus_table %>%
  dplyr::filter(.data$qc_status == "qc_pass") %>%
  dplyr::arrange(.data$gtex_chrom, .data$gene_name, .data$locus_id) %>%
  dplyr::mutate(
    task_id = dplyr::row_number(),
    output_dir = file.path(TASK_OUTPUT_ROOT, .data$locus_id)
  ) %>%
  dplyr::select(dplyr::all_of(c(
    "task_id", "locus_id", "gene_name", "gene_id", "gtex_tissue", "gtex_chrom",
    "n_variants", "n_rss_qc", "master_path", "order_path", "ld_long_path", "output_dir"
  )))

if (is.finite(TASK_LIMIT)) {
  task_manifest <- task_manifest %>% dplyr::slice_head(n = TASK_LIMIT)
}

readr::write_csv(task_manifest, TASK_MANIFEST_PATH)
cat("QC-passing task count:", nrow(task_manifest), "\n")
cat("Task manifest:", TASK_MANIFEST_PATH, "\n")
task_manifest %>% dplyr::slice_head(n = 10)
```

# 6. Write Slurm Script

```{r slurm-script, results='asis'}
if (!file.exists(RUN_TASK_SCRIPT)) {
  stop("Missing task script: ", RUN_TASK_SCRIPT)
}

task_count <- nrow(task_manifest)
if (task_count == 0L) {
  stop("No QC-passing loci available for baseline SuSiE screening.")
}

max_variants_arg <- if (is.finite(MAX_VARIANTS_FOR_SUSIE)) {
  as.character(MAX_VARIANTS_FOR_SUSIE)
} else {
  "Inf"
}
hpc_setup <- paste(
  c(
    if (isTRUE(LOAD_R_MODULE)) "module load r" else character(),
    if (nzchar(R_LIBS_USER)) sprintf('export R_LIBS_USER="%s"', R_LIBS_USER) else character()
  ),
  collapse = "\n"
)

slurm_text <- sprintf(
'#!/bin/bash
#SBATCH --job-name=%s
#SBATCH --array=1-%d
#SBATCH --time=%s
#SBATCH --mem=%s
#SBATCH --cpus-per-task=%d
#SBATCH --mail-user=%s
#SBATCH --mail-type=%s
#SBATCH --output=/dev/null
#SBATCH --error=/dev/null
#SBATCH --account=%s

set -euo pipefail

PROJECT_ROOT="%s"
TASK_MANIFEST="%s"
RUN_TASK_SCRIPT="%s"
SLURM_PRINTS_BASE="%s"

JOBNAME="${SLURM_JOB_NAME}"
PARENT_ID="${SLURM_ARRAY_JOB_ID:-$SLURM_JOB_ID}"
TASK_ID="${SLURM_ARRAY_TASK_ID:-1}"
PRINTS_DIR="${SLURM_PRINTS_BASE}/${JOBNAME}/${PARENT_ID}"
mkdir -p "${PRINTS_DIR}"

exec >"${PRINTS_DIR}/${TASK_ID}.out" 2>"${PRINTS_DIR}/${TASK_ID}.err"

echo "[$(date -Is)] Starting task ${TASK_ID} for job ${JOBNAME} (parent ${PARENT_ID})"
echo "Task manifest: ${TASK_MANIFEST}"

# --- site/module setup ---
%s

Rscript "${RUN_TASK_SCRIPT}" \\
  --task-manifest "${TASK_MANIFEST}" \\
  --task-id "${TASK_ID}" \\
  --L %d \\
  --max-variants %s \\
  --active-alpha-max-threshold %.6f

echo "[$(date -Is)] Completed task ${TASK_ID}"
',
  JOB_NAME,
  task_count,
  TIME_LIMIT,
  MEMORY,
  CPUS_PER_TASK,
  MAIL_USER,
  MAIL_TYPE,
  ACCOUNT,
  PROJECT_ROOT,
  TASK_MANIFEST_PATH,
  RUN_TASK_SCRIPT,
  SLURM_PRINTS_DIR,
  hpc_setup,
  SUSIE_L,
  max_variants_arg,
  ACTIVE_ALPHA_MAX_THRESHOLD
)

writeLines(slurm_text, SLURM_SCRIPT_PATH)
cat("Wrote Slurm script:", SLURM_SCRIPT_PATH, "\n\n")
cat("Submit with:\n\n")
cat("`sbatch ", SLURM_SCRIPT_PATH, "`\n", sep = "")
```

# 7. Smoke-Test Command

```{r smoke-command, results='asis'}
cat(
  "Local/HPC smoke command for the first task:\n\n",
  "`Rscript ", RUN_TASK_SCRIPT,
  " --task-manifest ", TASK_MANIFEST_PATH,
  " --task-id 1 --L ", SUSIE_L,
  " --max-variants ", max_variants_arg,
  " --active-alpha-max-threshold ", ACTIVE_ALPHA_MAX_THRESHOLD,
  "`\n",
  sep = ""
)
```
