---
title: "0.4 Explore Baseline SuSiE Metrics"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
editor_options:
  markdown:
    wrap: 72
---

This workbook collects compact vanilla SuSiE screening outputs, joins them to
Phase 1 QC/z/LD metrics, and renders exploratory plots for deciding a later
stratified resampling design.

```{r setup, message=FALSE}
required_packages <- c("dplyr", "ggplot2", "readr", "tibble", "tidyr", "purrr")
missing_packages <- required_packages[!vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing_packages) > 0) {
  stop(
    "Missing required R packages: ",
    paste(missing_packages, collapse = ", "),
    ". Install them before running this workbook."
  )
}

library(dplyr)
library(ggplot2)
library(readr)
library(tibble)
library(tidyr)
library(purrr)

theme_set(theme_bw(base_size = 11))
```

# 1. Configuration

```{r config}
find_project_root <- function(start = getwd()) {
  # Example fallback path -- set to your environment (or set EQTL_SUSINE_PROJECT_ROOT).
  hardcoded <- "/path/to/eQTL_annotations_for_susine"
  env_root <- Sys.getenv("EQTL_SUSINE_PROJECT_ROOT", unset = "")
  candidates <- unique(c(
    env_root,
    hardcoded,
    normalizePath(start, winslash = "/", mustWork = FALSE),
    dirname(normalizePath(start, winslash = "/", mustWork = FALSE)),
    dirname(dirname(normalizePath(start, winslash = "/", mustWork = FALSE)))
  ))
  candidates <- candidates[nzchar(candidates)]
  for (candidate in candidates) {
    if (
      dir.exists(file.path(candidate, "utils")) &&
      dir.exists(file.path(candidate, "vignettes")) &&
      file.exists(file.path(candidate, "README.md"))
    ) {
      return(normalizePath(candidate, winslash = "/", mustWork = TRUE))
    }
  }
  stop("Could not locate eQTL_annotations_for_susine project root.")
}

PROJECT_ROOT <- find_project_root()
MANIFEST_PATH <- file.path(PROJECT_ROOT, "config", "loci_manifest_sample_100_per_chrom.csv")
SCREEN_DIR <- file.path(PROJECT_ROOT, "output", "prelim", "baseline_susie_screen")
TASK_MANIFEST_PATH <- file.path(SCREEN_DIR, "baseline_susie_task_manifest.csv")
QC_TABLE_PATH <- file.path(SCREEN_DIR, "qc_locus_table.csv")
TASK_OUTPUT_ROOT <- file.path(PROJECT_ROOT, "output", "baseline_susie_screen", "task_outputs")
COLLECTED_DIR <- file.path(PROJECT_ROOT, "output", "baseline_susie_screen", "collected")
FIGURE_DIR <- file.path(PROJECT_ROOT, "output", "baseline_susie_screen", "figures")
GEOMETRIC_N20_DIR <- file.path(PROJECT_ROOT, "output", "baseline_susie_screen", "selected_geometric_n20")

dir.create(COLLECTED_DIR, recursive = TRUE, showWarnings = FALSE)
dir.create(FIGURE_DIR, recursive = TRUE, showWarnings = FALSE)
dir.create(GEOMETRIC_N20_DIR, recursive = TRUE, showWarnings = FALSE)

LOCUS_METRICS_PATH <- file.path(COLLECTED_DIR, "baseline_susie_locus_metrics.csv")
EFFECT_METRICS_PATH <- file.path(COLLECTED_DIR, "baseline_susie_effect_metrics.csv")
TOP_VARIANTS_PATH <- file.path(COLLECTED_DIR, "baseline_susie_top_variants.csv")
JOINED_METRICS_PATH <- file.path(COLLECTED_DIR, "baseline_susie_metric_joined.csv")
MISSING_TASKS_PATH <- file.path(COLLECTED_DIR, "baseline_susie_missing_task_outputs.csv")
LOCKED_GEOMETRIC_N20_PATH <- file.path(GEOMETRIC_N20_DIR, "locked_loci.csv")
LOCKED_GEOMETRIC_N20_MANIFEST_PATH <- file.path(GEOMETRIC_N20_DIR, "locked_loci_with_manifest.csv")
LOCKED_GEOMETRIC_N20_SELECTION_PATH <- file.path(GEOMETRIC_N20_DIR, "annotation_selection_geometric_n20.csv")
CONFIG_GEOMETRIC_N20_SELECTION_PATH <- file.path(PROJECT_ROOT, "config", "annotation_selection_geometric_n20.csv")

LABEL_LOCI <- c("mei1_chr22_lung", "pus7l_chr12_lung")
Z_HIGH_THRESHOLD <- 5
PIP_HIGH_THRESHOLD <- 0.5
PIP_STRONG_THRESHOLD <- 0.9

cat("Project root:", PROJECT_ROOT, "\n")
cat("Screen dir:", SCREEN_DIR, "\n")
cat("Task output root:", TASK_OUTPUT_ROOT, "\n")
```

# 2. Collection Helpers

```{r helpers}
task_output_col_types <- function(filename) {
  if (identical(filename, "locus_metrics.csv")) {
    return(readr::cols(
      locus_id = readr::col_character(),
      gene_name = readr::col_character(),
      n_variants = readr::col_integer(),
      n_rss = readr::col_integer(),
      status = readr::col_character(),
      error_message = readr::col_character(),
      elapsed_seconds = readr::col_double(),
      elbo_last = readr::col_double(),
      n_iter = readr::col_integer(),
      max_pip = readr::col_double(),
      pip_sum = readr::col_double(),
      pip_entropy = readr::col_double(),
      pip_k_eff = readr::col_double(),
      pip_top1_share = readr::col_double(),
      pip_top2_gap = readr::col_double(),
      n_pip_gt_0_01 = readr::col_integer(),
      n_pip_gt_0_05 = readr::col_integer(),
      n_pip_gt_0_1 = readr::col_integer(),
      n_pip_gt_0_5 = readr::col_integer(),
      n_pip_gt_0_9 = readr::col_integer(),
      z_max_abs = readr::col_double(),
      z_topk_ratio = readr::col_double(),
      z_count_abs_gt_3 = readr::col_integer(),
      z_eff_signals = readr::col_double(),
      n_active_effects = readr::col_integer(),
      min_active_alpha_max = readr::col_double(),
      median_active_alpha_max = readr::col_double(),
      max_active_alpha_max = readr::col_double(),
      min_active_alpha_k_eff_core95 = readr::col_double(),
      median_active_alpha_k_eff_core95 = readr::col_double(),
      max_active_alpha_k_eff_core95 = readr::col_double(),
      .default = readr::col_guess()
    ))
  }
  if (identical(filename, "effect_metrics.csv")) {
    return(readr::cols(
      locus_id = readr::col_character(),
      gene_name = readr::col_character(),
      effect_l = readr::col_integer(),
      alpha_max = readr::col_double(),
      alpha_entropy = readr::col_double(),
      alpha_entropy_core95 = readr::col_double(),
      alpha_k_eff = readr::col_double(),
      alpha_k_eff_core95 = readr::col_double(),
      cs_size_raw = readr::col_integer(),
      cs_purity = readr::col_double(),
      .default = readr::col_guess()
    ))
  }
  if (identical(filename, "top_variants.csv")) {
    return(readr::cols(
      locus_id = readr::col_character(),
      gene_name = readr::col_character(),
      rank_pip = readr::col_integer(),
      variant_id = readr::col_character(),
      ld_matrix_index = readr::col_integer(),
      chrom = readr::col_character(),
      pos = readr::col_integer(),
      ref = readr::col_character(),
      alt = readr::col_character(),
      z_score = readr::col_double(),
      pip = readr::col_double(),
      posterior_mean = readr::col_double(),
      .default = readr::col_guess()
    ))
  }
  readr::cols(.default = readr::col_guess())
}

read_optional_csv <- function(path, filename = basename(path)) {
  if (!file.exists(path)) return(NULL)
  readr::read_csv(path, col_types = task_output_col_types(filename), show_col_types = FALSE)
}

bind_task_csv <- function(task_manifest, filename) {
  rows <- purrr::map(seq_len(nrow(task_manifest)), function(i) {
    path <- file.path(task_manifest$output_dir[[i]], filename)
    tbl <- read_optional_csv(path, filename = filename)
    if (is.null(tbl)) return(NULL)
    tbl
  })
  rows <- rows[!vapply(rows, is.null, logical(1))]
  if (!length(rows)) tibble() else dplyr::bind_rows(rows)
}

empty_locus_metrics <- function() {
  tibble(
    locus_id = character(),
    gene_name = character(),
    status = character(),
    n_variants = integer(),
    n_rss = integer(),
    elapsed_seconds = double(),
    elbo_last = double(),
    n_iter = integer(),
    max_pip = double(),
    pip_sum = double(),
    pip_entropy = double(),
    pip_k_eff = double(),
    pip_top1_share = double(),
    pip_top2_gap = double(),
    n_pip_gt_0_01 = integer(),
    n_pip_gt_0_05 = integer(),
    n_pip_gt_0_1 = integer(),
    n_pip_gt_0_5 = integer(),
    n_pip_gt_0_9 = integer(),
    z_max_abs = double(),
    z_topk_ratio = double(),
    z_count_abs_gt_3 = integer(),
    z_eff_signals = double(),
    n_active_effects = integer(),
    min_active_alpha_max = double(),
    median_active_alpha_max = double(),
    max_active_alpha_max = double(),
    min_active_alpha_k_eff_core95 = double(),
    median_active_alpha_k_eff_core95 = double(),
    max_active_alpha_k_eff_core95 = double()
  )
}

empty_effect_metrics <- function() {
  tibble(
    locus_id = character(),
    gene_name = character(),
    effect_l = integer(),
    alpha_max = double(),
    alpha_entropy = double(),
    alpha_entropy_core95 = double(),
    alpha_k_eff = double(),
    alpha_k_eff_core95 = double(),
    cs_size_raw = integer(),
    cs_purity = double()
  )
}

empty_top_variants <- function() {
  tibble(
    locus_id = character(),
    gene_name = character(),
    rank_pip = integer(),
    variant_id = character(),
    ld_matrix_index = integer(),
    chrom = character(),
    pos = integer(),
    ref = character(),
    alt = character(),
    z_score = double(),
    pip = double(),
    posterior_mean = double()
  )
}

numeric_cols <- function(df) {
  names(df)[vapply(df, is.numeric, logical(1))]
}

safe_cor <- function(x, y) {
  keep <- is.finite(x) & is.finite(y)
  if (sum(keep) < 3L) return(NA_real_)
  suppressWarnings(stats::cor(x[keep], y[keep], method = "spearman"))
}

robust_scale <- function(df) {
  out <- df
  for (nm in names(out)) {
    x <- as.numeric(out[[nm]])
    med <- stats::median(x, na.rm = TRUE)
    iqr <- stats::IQR(x, na.rm = TRUE)
    if (!is.finite(iqr) || iqr == 0) {
      iqr <- stats::sd(x, na.rm = TRUE)
    }
    if (!is.finite(iqr) || iqr == 0) {
      iqr <- 1
    }
    out[[nm]] <- (x - med) / iqr
  }
  out
}

complete_pair_count <- function(x, y) {
  sum(is.finite(x) & is.finite(y))
}

first_existing_col <- function(df, candidates) {
  hits <- candidates[candidates %in% names(df)]
  if (length(hits)) hits[[1]] else NA_character_
}

copy_first_available <- function(df, target, candidates) {
  source_col <- first_existing_col(df, candidates)
  if (is.na(source_col)) {
    df[[target]] <- NA_real_
  } else {
    df[[target]] <- df[[source_col]]
  }
  df
}

transform_for_pca <- function(df, cols) {
  x <- df %>% dplyr::select(dplyr::all_of(cols))
  for (nm in names(x)) {
    if (grepl("count|n_|k_eff|eff_signals", nm)) {
      vals <- as.numeric(x[[nm]])
      min_val <- suppressWarnings(min(vals, na.rm = TRUE))
      if (is.finite(min_val) && min_val >= 0) {
        x[[nm]] <- log1p(vals)
      }
    }
  }
  x
}

drop_bad_pca_cols <- function(df, cols, min_complete = 3L) {
  cols[vapply(cols, function(nm) {
    x <- as.numeric(df[[nm]])
    keep <- is.finite(x)
    sum(keep) >= min_complete && is.finite(stats::sd(x[keep])) && stats::sd(x[keep]) > 0
  }, logical(1))]
}

make_cv_folds <- function(n, k = 5L, seed = 1L) {
  set.seed(seed)
  sample(rep(seq_len(k), length.out = n))
}

label_loci_layer <- function(df, x_col, y_col) {
  label_df <- df %>% dplyr::filter(.data$locus_id %in% LABEL_LOCI)
  if (!nrow(label_df)) return(NULL)
  geom_text(
    data = label_df,
    aes(x = .data[[x_col]], y = .data[[y_col]], label = .data$locus_id),
    size = 3,
    nudge_y = 0.03,
    check_overlap = TRUE
  )
}
```

# 3. Collect Outputs

```{r collect}
task_manifest <- readr::read_csv(TASK_MANIFEST_PATH, show_col_types = FALSE)
qc_locus_table <- readr::read_csv(QC_TABLE_PATH, show_col_types = FALSE)

missing_task_outputs <- task_manifest %>%
  dplyr::mutate(
    locus_metrics_path = file.path(.data$output_dir, "locus_metrics.csv"),
    effect_metrics_path = file.path(.data$output_dir, "effect_metrics.csv"),
    top_variants_path = file.path(.data$output_dir, "top_variants.csv"),
    has_locus_metrics = file.exists(.data$locus_metrics_path),
    has_effect_metrics = file.exists(.data$effect_metrics_path),
    has_top_variants = file.exists(.data$top_variants_path)
  ) %>%
  dplyr::filter(!.data$has_locus_metrics | !.data$has_effect_metrics | !.data$has_top_variants)

readr::write_csv(missing_task_outputs, MISSING_TASKS_PATH)

locus_metrics <- dplyr::bind_rows(empty_locus_metrics(), bind_task_csv(task_manifest, "locus_metrics.csv"))
effect_metrics <- dplyr::bind_rows(empty_effect_metrics(), bind_task_csv(task_manifest, "effect_metrics.csv"))
top_variants <- dplyr::bind_rows(empty_top_variants(), bind_task_csv(task_manifest, "top_variants.csv"))

joined_metrics <- qc_locus_table %>%
  dplyr::left_join(
    locus_metrics,
    by = c("locus_id", "gene_name"),
    suffix = c("_phase1", "_susie")
  ) %>%
  dplyr::mutate(
    phase1_status = dplyr::coalesce(.data$status_phase1, NA_character_),
    susie_status = dplyr::coalesce(.data$status_susie, NA_character_),
    status = .data$susie_status,
    z_signal_zone = dplyr::case_when(
      is.finite(.data$z_max_abs) & .data$z_max_abs >= Z_HIGH_THRESHOLD ~ "high_z",
      is.finite(.data$z_max_abs) ~ "low_or_moderate_z",
      TRUE ~ "missing_z"
    ),
    pip_zone = dplyr::case_when(
      is.finite(.data$max_pip) & .data$max_pip >= PIP_STRONG_THRESHOLD ~ "pip_ge_0p9",
      is.finite(.data$max_pip) & .data$max_pip >= PIP_HIGH_THRESHOLD ~ "pip_0p5_to_0p9",
      is.finite(.data$max_pip) ~ "pip_lt_0p5",
      TRUE ~ "missing_pip"
    ),
    signal_pip_zone = paste(.data$z_signal_zone, .data$pip_zone, sep = "__")
  )

joined_metrics <- joined_metrics %>%
  copy_first_available("z_eff_signals_for_sampling", c("z_eff_signals_phase1", "z_eff_signals_susie", "z_eff_signals")) %>%
  copy_first_available("z_count_abs_gt_3_for_sampling", c("z_count_abs_gt_3_phase1", "z_count_abs_gt_3_susie", "z_count_abs_gt_3")) %>%
  copy_first_available("n_variants_for_sampling", c("n_variants_phase1", "n_variants_susie", "phase1_master_variants", "ld_matched_variants"))

readr::write_csv(locus_metrics, LOCUS_METRICS_PATH)
readr::write_csv(effect_metrics, EFFECT_METRICS_PATH)
readr::write_csv(top_variants, TOP_VARIANTS_PATH)
readr::write_csv(joined_metrics, JOINED_METRICS_PATH)

signal_metric_cols <- intersect(
  c(
    "z_max_abs", "z_topk_ratio", "z_count_abs_gt_3_susie", "z_eff_signals_susie",
    "z_count_abs_gt_3_phase1", "z_eff_signals_phase1",
    "z_count_abs_gt_3_for_sampling", "z_eff_signals_for_sampling"
  ),
  names(joined_metrics)
)
posterior_metric_cols <- intersect(
  c(
    "max_pip", "pip_sum", "pip_entropy", "pip_k_eff", "pip_top1_share",
    "pip_top2_gap", "n_pip_gt_0_01", "n_pip_gt_0_05", "n_pip_gt_0_1",
    "n_pip_gt_0_5", "n_pip_gt_0_9"
  ),
  names(joined_metrics)
)
effect_summary_metric_cols <- intersect(
  c(
    "n_active_effects", "min_active_alpha_max", "median_active_alpha_max",
    "max_active_alpha_max", "min_active_alpha_k_eff_core95",
    "median_active_alpha_k_eff_core95", "max_active_alpha_k_eff_core95"
  ),
  names(joined_metrics)
)
run_bookkeeping_cols <- intersect(
  c("task_id", "elapsed_seconds", "elbo_last", "n_iter", "n_rss", "n_rss_qc"),
  names(joined_metrics)
)
susie_metric_cols <- unique(c(signal_metric_cols, posterior_metric_cols, effect_summary_metric_cols))
dataset_difficulty_metric_cols <- intersect(
  c(
    "M1", "M2", "M4",
    "raw_gene_variants_loaded", "post_af_filter", "post_sample_size_filter",
    "z_score_variants_exported", "phase1_master_variants",
    "ld_input_variants", "ld_matched_variants", "ld_unmatched_variants",
    "matched_variants_in_1000g", "n_variants_phase1", "n_variants_for_sampling",
    "z_eff_signals_phase1", "z_count_abs_gt_3_phase1",
    "z_eff_signals_for_sampling", "z_count_abs_gt_3_for_sampling"
  ),
  setdiff(names(joined_metrics), c(run_bookkeeping_cols, posterior_metric_cols, effect_summary_metric_cols))
)

metric_inventory <- tibble(
  metric = unique(c(dataset_difficulty_metric_cols, signal_metric_cols, posterior_metric_cols, effect_summary_metric_cols))
) %>%
  dplyr::mutate(
    metric_group = dplyr::case_when(
    .data$metric %in% dataset_difficulty_metric_cols ~ "phase1_qc_dataset_difficulty",
    .data$metric %in% signal_metric_cols ~ "aligned_z_signal",
    .data$metric %in% posterior_metric_cols ~ "baseline_susie_posterior",
    .data$metric %in% effect_summary_metric_cols ~ "baseline_susie_effect_summary",
    TRUE ~ "other"
    )
  )
readr::write_csv(metric_inventory, file.path(COLLECTED_DIR, "baseline_susie_metric_inventory.csv"))

list(
  task_count = nrow(task_manifest),
  missing_task_outputs = nrow(missing_task_outputs),
  collected_locus_rows = nrow(locus_metrics),
  collected_effect_rows = nrow(effect_metrics),
  collected_top_variant_rows = nrow(top_variants),
  dataset_difficulty_metrics = length(dataset_difficulty_metric_cols),
  susie_signal_posterior_effect_metrics = length(susie_metric_cols)
)
```

```{r status-summary}
joined_metrics %>%
  dplyr::count(.data$qc_status, .data$status, .data$signal_pip_zone, sort = TRUE)
```

# 4. Histograms

```{r histograms, fig.width=16, fig.height=12}
hist_metric_cols <- setdiff(numeric_cols(joined_metrics), c("task_id"))
hist_df <- joined_metrics %>%
  dplyr::select(dplyr::all_of(c("locus_id", hist_metric_cols))) %>%
  tidyr::pivot_longer(-.data$locus_id, names_to = "metric", values_to = "value") %>%
  dplyr::filter(is.finite(.data$value))

if (nrow(hist_df) > 0) {
  p_hist <- ggplot(hist_df, aes(x = .data$value)) +
    geom_histogram(bins = 35, fill = "#33658A", color = "white", linewidth = 0.2) +
    facet_wrap(~ metric, scales = "free", ncol = 5) +
    labs(title = "Histograms for Phase 1, QC, z, PIP, and effect-diffuseness metrics",
         x = NULL, y = "Locus count")
} else {
  p_hist <- ggplot() +
    annotate("text", x = 0, y = 0, label = "No finite numeric metrics available yet.") +
    theme_void()
}

ggsave(file.path(FIGURE_DIR, "all_metric_histograms.png"), p_hist, width = 16, height = 12, dpi = 150)
p_hist
```

# 5. Correlations and Scatterplots

```{r correlation-heatmap, fig.width=10, fig.height=8}
focus_metrics <- c(
  "z_max_abs", "z_topk_ratio", "z_eff_signals_for_sampling",
  "max_pip", "pip_k_eff", "pip_top2_gap",
  "M1", "M2", "M4",
  "n_active_effects", "median_active_alpha_k_eff_core95"
)
focus_metrics <- intersect(focus_metrics, names(joined_metrics))

cor_pairs <- tidyr::expand_grid(x_metric = focus_metrics, y_metric = focus_metrics) %>%
  dplyr::mutate(
    rho = purrr::map2_dbl(
      .data$x_metric,
      .data$y_metric,
      ~ safe_cor(joined_metrics[[.x]], joined_metrics[[.y]])
    )
  )

p_cor <- ggplot(cor_pairs, aes(x = .data$x_metric, y = .data$y_metric, fill = .data$rho)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "#2166AC", mid = "white", high = "#B2182B", limits = c(-1, 1), na.value = "grey85") +
  coord_equal() +
  labs(title = "Spearman correlation heatmap", x = NULL, y = NULL, fill = "rho") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave(file.path(FIGURE_DIR, "metric_correlation_heatmap.png"), p_cor, width = 10, height = 8, dpi = 150)
p_cor
```

```{r difficulty-vs-susie-correlations, fig.width=14, fig.height=10}
difficulty_vs_susie_pairs <- tidyr::expand_grid(
  difficulty_metric = dataset_difficulty_metric_cols,
  susie_metric = susie_metric_cols
) %>%
  dplyr::mutate(
    n_complete = purrr::map2_int(
      .data$difficulty_metric,
      .data$susie_metric,
      ~ complete_pair_count(joined_metrics[[.x]], joined_metrics[[.y]])
    ),
    rho = purrr::map2_dbl(
      .data$difficulty_metric,
      .data$susie_metric,
      ~ safe_cor(joined_metrics[[.x]], joined_metrics[[.y]])
    )
  ) %>%
  dplyr::arrange(dplyr::desc(abs(.data$rho)))

readr::write_csv(
  difficulty_vs_susie_pairs,
  file.path(COLLECTED_DIR, "baseline_susie_difficulty_vs_susie_correlations.csv")
)

if (nrow(difficulty_vs_susie_pairs) > 0) {
  p_diff_susie_cor <- ggplot(
    difficulty_vs_susie_pairs,
    aes(x = .data$susie_metric, y = .data$difficulty_metric, fill = .data$rho)
  ) +
    geom_tile(color = "white") +
    scale_fill_gradient2(low = "#2166AC", mid = "white", high = "#B2182B", limits = c(-1, 1), na.value = "grey85") +
    labs(
      title = "Dataset difficulty metrics vs baseline SuSiE metrics",
      x = "Baseline SuSiE / aligned-z metric",
      y = "Phase-1/QC dataset difficulty metric",
      fill = "Spearman rho"
    ) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
} else {
  p_diff_susie_cor <- ggplot() +
    annotate("text", x = 0, y = 0, label = "No dataset difficulty and SuSiE metric pairs available yet.") +
    theme_void()
}

ggsave(file.path(FIGURE_DIR, "difficulty_vs_susie_correlation_heatmap.png"), p_diff_susie_cor, width = 14, height = 10, dpi = 150)
p_diff_susie_cor
```

```{r scatterplots, fig.width=14, fig.height=10}
scatter_pairs <- tibble::tribble(
  ~x_metric, ~y_metric,
  "z_max_abs", "max_pip",
  "z_topk_ratio", "max_pip",
  "z_eff_signals_for_sampling", "pip_k_eff",
  "z_eff_signals_for_sampling", "max_pip",
  "M1", "max_pip",
  "M1", "pip_k_eff",
  "M2", "max_pip",
  "M4", "max_pip",
  "pip_top2_gap", "max_pip",
  "median_active_alpha_k_eff_core95", "max_pip"
) %>%
  dplyr::filter(.data$x_metric %in% names(joined_metrics), .data$y_metric %in% names(joined_metrics))

scatter_df <- purrr::pmap_dfr(scatter_pairs, function(x_metric, y_metric) {
  joined_metrics %>%
    dplyr::transmute(
      locus_id = .data$locus_id,
      qc_status = .data$qc_status,
      signal_pip_zone = .data$signal_pip_zone,
      x_value = .data[[x_metric]],
      y_value = .data[[y_metric]],
      pair = paste(x_metric, "vs", y_metric)
    )
})

if (nrow(scatter_df) > 0) {
  p_scatter <- ggplot(scatter_df, aes(x = .data$x_value, y = .data$y_value, color = .data$signal_pip_zone)) +
    geom_point(alpha = 0.75, size = 1.8) +
    facet_wrap(~ pair, scales = "free", ncol = 3) +
    labs(title = "Focused metric scatterplots", x = NULL, y = NULL, color = "Zone") +
    theme(legend.position = "bottom")
} else {
  p_scatter <- ggplot() +
    annotate("text", x = 0, y = 0, label = "No focused metric pairs available yet.") +
    theme_void()
}

ggsave(file.path(FIGURE_DIR, "metric_pairwise_scatter.png"), p_scatter, width = 14, height = 10, dpi = 150)
p_scatter
```

```{r difficulty-vs-susie-top-scatterplots, fig.width=16, fig.height=12}
top_difficulty_susie_pairs <- difficulty_vs_susie_pairs %>%
  dplyr::filter(is.finite(.data$rho), .data$n_complete >= 3L) %>%
  dplyr::slice_head(n = 24)

top_difficulty_susie_scatter_df <- purrr::pmap_dfr(
  top_difficulty_susie_pairs,
  function(difficulty_metric, susie_metric, n_complete, rho) {
    joined_metrics %>%
      dplyr::transmute(
        locus_id = .data$locus_id,
        signal_pip_zone = .data$signal_pip_zone,
        x_value = .data[[difficulty_metric]],
        y_value = .data[[susie_metric]],
        pair = sprintf("%s vs %s (rho=%.2f)", difficulty_metric, susie_metric, rho)
      )
  }
)

if (nrow(top_difficulty_susie_scatter_df) > 0) {
  p_top_diff_scatter <- ggplot(
    top_difficulty_susie_scatter_df,
    aes(x = .data$x_value, y = .data$y_value, color = .data$signal_pip_zone)
  ) +
    geom_point(alpha = 0.75, size = 1.8) +
    facet_wrap(~ pair, scales = "free", ncol = 4) +
    labs(
      title = "Top dataset difficulty vs baseline SuSiE metric relationships",
      x = NULL,
      y = NULL,
      color = "Zone"
    ) +
    theme(legend.position = "bottom")
} else {
  p_top_diff_scatter <- ggplot() +
    annotate("text", x = 0, y = 0, label = "No finite difficulty-vs-SuSiE scatter pairs available yet.") +
    theme_void()
}

ggsave(file.path(FIGURE_DIR, "top_difficulty_vs_susie_scatter.png"), p_top_diff_scatter, width = 16, height = 12, dpi = 150)
p_top_diff_scatter
```

# 6. PCA on Signal and Posterior Metrics

```{r pca, fig.width=9, fig.height=7}
pca_metrics <- c(
  "z_max_abs", "z_topk_ratio", "z_eff_signals_for_sampling",
  "max_pip", "pip_k_eff", "pip_top2_gap",
  "n_active_effects", "median_active_alpha_k_eff_core95",
  "max_active_alpha_k_eff_core95"
)
pca_metrics <- intersect(pca_metrics, names(joined_metrics))

pca_input <- joined_metrics %>%
  dplyr::filter(.data$qc_status == "qc_pass", .data$status == "completed") %>%
  dplyr::select(dplyr::all_of(c("locus_id", "signal_pip_zone", pca_metrics))) %>%
  tidyr::drop_na(dplyr::all_of(pca_metrics))

if (nrow(pca_input) >= 3L && length(pca_metrics) >= 2L) {
  x <- transform_for_pca(pca_input, pca_metrics) %>%
    dplyr::select(dplyr::all_of(pca_metrics)) %>%
    robust_scale()
  pca_fit <- stats::prcomp(x, center = FALSE, scale. = FALSE)
  pca_scores <- pca_input %>%
    dplyr::select(dplyr::all_of(c("locus_id", "signal_pip_zone"))) %>%
    dplyr::bind_cols(as_tibble(pca_fit$x[, seq_len(min(5L, ncol(pca_fit$x))), drop = FALSE]))
  pca_loadings <- as_tibble(pca_fit$rotation[, seq_len(min(5L, ncol(pca_fit$rotation))), drop = FALSE], rownames = "metric")
  readr::write_csv(pca_scores, file.path(COLLECTED_DIR, "baseline_susie_signal_posterior_pca_scores.csv"))
  readr::write_csv(pca_loadings, file.path(COLLECTED_DIR, "baseline_susie_signal_posterior_pca_loadings.csv"))

  p_pca <- ggplot(pca_scores, aes(x = .data$PC1, y = .data$PC2, color = .data$signal_pip_zone)) +
    geom_point(size = 2.2, alpha = 0.8) +
    geom_text(
      data = pca_scores %>% dplyr::filter(.data$locus_id %in% LABEL_LOCI),
      aes(label = .data$locus_id),
      nudge_y = 0.05,
      size = 3,
      check_overlap = TRUE
    ) +
    labs(
      title = "PCA on z and vanilla SuSiE posterior metrics",
      x = "PC1",
      y = "PC2",
      color = "Zone"
    ) +
    theme(legend.position = "bottom")

  ggsave(file.path(FIGURE_DIR, "signal_posterior_pca.png"), p_pca, width = 9, height = 7, dpi = 150)
  p_pca
} else {
  message("Skipping PCA: need at least 3 complete rows and 2 metrics.")
}
```

# 7. Joint PCA on Difficulty and SuSiE Metrics

```{r joint-difficulty-susie-pca, fig.width=10, fig.height=8}
joint_pca_metrics <- unique(c(dataset_difficulty_metric_cols, susie_metric_cols))
joint_pca_metrics <- intersect(joint_pca_metrics, names(joined_metrics))
joint_pca_metrics <- drop_bad_pca_cols(joined_metrics, joint_pca_metrics)

joint_pca_input <- joined_metrics %>%
  dplyr::filter(.data$qc_status == "qc_pass", .data$status == "completed") %>%
  dplyr::select(dplyr::all_of(c("locus_id", "signal_pip_zone", joint_pca_metrics))) %>%
  tidyr::drop_na(dplyr::all_of(joint_pca_metrics))

if (nrow(joint_pca_input) >= 3L && length(joint_pca_metrics) >= 2L) {
  x_joint <- transform_for_pca(joint_pca_input, joint_pca_metrics) %>%
    dplyr::select(dplyr::all_of(joint_pca_metrics)) %>%
    robust_scale()
  joint_pca_fit <- stats::prcomp(x_joint, center = FALSE, scale. = FALSE)
  joint_pca_scores <- joint_pca_input %>%
    dplyr::select(dplyr::all_of(c("locus_id", "signal_pip_zone"))) %>%
    dplyr::bind_cols(as_tibble(joint_pca_fit$x[, seq_len(min(5L, ncol(joint_pca_fit$x))), drop = FALSE]))
  joint_pca_loadings <- as_tibble(
    joint_pca_fit$rotation[, seq_len(min(5L, ncol(joint_pca_fit$rotation))), drop = FALSE],
    rownames = "metric"
  ) %>%
    dplyr::left_join(metric_inventory, by = "metric")

  readr::write_csv(joint_pca_scores, file.path(COLLECTED_DIR, "baseline_susie_joint_difficulty_pca_scores.csv"))
  readr::write_csv(joint_pca_loadings, file.path(COLLECTED_DIR, "baseline_susie_joint_difficulty_pca_loadings.csv"))

  p_joint_pca <- ggplot(joint_pca_scores, aes(x = .data$PC1, y = .data$PC2, color = .data$signal_pip_zone)) +
    geom_point(size = 2.2, alpha = 0.8) +
    geom_text(
      data = joint_pca_scores %>% dplyr::filter(.data$locus_id %in% LABEL_LOCI),
      aes(label = .data$locus_id),
      nudge_y = 0.05,
      size = 3,
      check_overlap = TRUE
    ) +
    labs(
      title = "Joint PCA on dataset difficulty and baseline SuSiE metrics",
      x = "PC1",
      y = "PC2",
      color = "Zone"
    ) +
    theme(legend.position = "bottom")

  ggsave(file.path(FIGURE_DIR, "joint_difficulty_susie_pca.png"), p_joint_pca, width = 10, height = 8, dpi = 150)
  p_joint_pca
} else {
  message("Skipping joint PCA: need at least 3 complete rows and 2 usable metrics.")
}
```

# 8. Predict Posterior Diffuseness From Dataset Metrics

This section predicts baseline SuSiE posterior diffuseness from pre/post-QC
dataset metrics. The stratification target is not the prediction alone; it is
the predicted-vs-actual position and residual. Positive residuals are loci where
baseline SuSiE is more diffuse than the dataset metrics would suggest.

```{r predict-pip-k-eff, fig.width=8, fig.height=6}
PIP_KEFF_PREDICTION_PATH <- file.path(COLLECTED_DIR, "baseline_susie_pip_k_eff_prediction.csv")
PIP_KEFF_PREDICTION_LOADINGS_PATH <- file.path(COLLECTED_DIR, "baseline_susie_pip_k_eff_prediction_pca_loadings.csv")
PIP_KEFF_PREDICTION_SUMMARY_PATH <- file.path(COLLECTED_DIR, "baseline_susie_pip_k_eff_prediction_summary.csv")
PIP_KEFF_RESIDUAL_CANDIDATES_PATH <- file.path(COLLECTED_DIR, "baseline_susie_pip_k_eff_residual_candidates.csv")
PIP_KEFF_GEOMETRIC_SAMPLE_PATH <- file.path(COLLECTED_DIR, "baseline_susie_pip_k_eff_geometric_sample_n20.csv")
PIP_KEFF_GEOMETRIC_SAMPLE_COUNTS_PATH <- file.path(COLLECTED_DIR, "baseline_susie_pip_k_eff_geometric_sample_region_counts.csv")
PIP_KEFF_RESIDUAL_GEOMETRIC_SAMPLE_PATH <- file.path(COLLECTED_DIR, "baseline_susie_pip_k_eff_residual_geometry_sample_n20.csv")
PIP_KEFF_RESIDUAL_GEOMETRIC_SAMPLE_COUNTS_PATH <- file.path(COLLECTED_DIR, "baseline_susie_pip_k_eff_residual_geometry_sample_region_counts.csv")
PIP_KEFF_GEOMETRIC_SAMPLE_COMPARISON_COUNTS_PATH <- file.path(COLLECTED_DIR, "baseline_susie_pip_k_eff_geometric_sample_comparison_counts.csv")

GEOMETRIC_SAMPLE_N <- 20L
GEOMETRIC_FORCED_LOCI <- intersect("pus7l_chr12_lung", LABEL_LOCI)

scale_geometry_coords <- function(df, cols) {
  x <- as.matrix(df[, cols, drop = FALSE])
  center <- apply(x, 2, stats::median, na.rm = TRUE)
  scale <- apply(x, 2, stats::IQR, na.rm = TRUE)
  sd_scale <- apply(x, 2, stats::sd, na.rm = TRUE)
  scale[!is.finite(scale) | scale == 0] <- sd_scale[!is.finite(scale) | scale == 0]
  scale[!is.finite(scale) | scale == 0] <- 1
  sweep(sweep(x, 2, center, "-"), 2, scale, "/")
}

repulsive_geometric_sample <- function(df, x_col, y_col, n, forced_loci = character()) {
  geom_df <- df %>%
    dplyr::filter(is.finite(.data[[x_col]]), is.finite(.data[[y_col]]))
  if (!nrow(geom_df)) return(geom_df)
  n <- min(as.integer(n), nrow(geom_df))
  coords <- scale_geometry_coords(geom_df, c(x_col, y_col))
  forced_idx <- which(geom_df$locus_id %in% forced_loci)
  selected <- unique(forced_idx)

  if (!length(selected)) {
    selected <- which.min(rowSums(coords^2))
  }

  while (length(selected) < n) {
    candidate_idx <- setdiff(seq_len(nrow(geom_df)), selected)
    d_to_selected <- sapply(candidate_idx, function(i) {
      min(sqrt(rowSums((t(t(coords[selected, , drop = FALSE]) - coords[i, ]))^2)))
    })
    selected <- c(selected, candidate_idx[which.max(d_to_selected)])
  }

  geom_df[selected, , drop = FALSE] %>%
    dplyr::mutate(geometric_sample_order = seq_len(dplyr::n()))
}

assign_interpretive_region <- function(df) {
  abs_resid_cut <- stats::quantile(abs(df$log_pip_k_eff_residual), 0.4, na.rm = TRUE)
  df %>%
    dplyr::mutate(
      suggested_sampling_region = dplyr::case_when(
        .data$predicted_pip_k_eff_bin == 3L &
          .data$actual_pip_k_eff_bin == 3L &
          abs(.data$log_pip_k_eff_residual) <= abs_resid_cut ~ "expected_diffuse",
        .data$residual_zone == "more_diffuse_than_predicted" ~ "surprisingly_diffuse",
        .data$residual_zone == "less_diffuse_than_predicted" &
          .data$predicted_pip_k_eff_bin == 3L ~ "surprisingly_concentrated",
        .data$predicted_pip_k_eff_bin == 1L &
          .data$actual_pip_k_eff_bin == 1L ~ "easy_control",
        .data$predicted_pip_k_eff_bin == 1L &
          .data$actual_pip_k_eff_bin == 3L ~ "low_predicted_diffuse_outlier",
        TRUE ~ "typical_or_transition"
      )
    )
}

target_col <- "pip_k_eff"
pip_keff_predictor_cols <- unique(c(
  "M1", "M2", "M4",
  "z_eff_signals_for_sampling", "z_count_abs_gt_3_for_sampling",
  "n_variants_for_sampling", "phase1_master_variants",
  "ld_matched_variants", "ld_unmatched_variants", "matched_variants_in_1000g",
  "raw_gene_variants_loaded", "post_af_filter", "post_sample_size_filter",
  "z_score_variants_exported"
))
pip_keff_predictor_cols <- intersect(pip_keff_predictor_cols, dataset_difficulty_metric_cols)
pip_keff_predictor_cols <- drop_bad_pca_cols(joined_metrics, pip_keff_predictor_cols, min_complete = 10L)

pip_keff_model_df <- joined_metrics %>%
  dplyr::filter(.data$qc_status == "qc_pass", .data$status == "completed") %>%
  dplyr::filter(is.finite(.data[[target_col]]), .data[[target_col]] > 0) %>%
  dplyr::select(dplyr::all_of(c(
    "locus_id", "gene_name", "signal_pip_zone", "max_pip",
    target_col, pip_keff_predictor_cols
  ))) %>%
  tidyr::drop_na(dplyr::all_of(c(target_col, pip_keff_predictor_cols)))

if (nrow(pip_keff_model_df) >= 20L && length(pip_keff_predictor_cols) >= 2L) {
  transformed_predictors <- transform_for_pca(pip_keff_model_df, pip_keff_predictor_cols) %>%
    dplyr::select(dplyr::all_of(pip_keff_predictor_cols))
  folds <- make_cv_folds(nrow(pip_keff_model_df), k = 5L, seed = 20260509L)
  target_log <- log1p(pip_keff_model_df[[target_col]])
  cv_pred_log <- rep(NA_real_, nrow(pip_keff_model_df))
  cv_n_pcs <- rep(NA_integer_, nrow(pip_keff_model_df))

  for (fold in sort(unique(folds))) {
    train_idx <- which(folds != fold)
    test_idx <- which(folds == fold)
    train_x <- transformed_predictors[train_idx, , drop = FALSE] %>% robust_scale()
    test_x_raw <- transformed_predictors[test_idx, , drop = FALSE]
    train_center <- vapply(transformed_predictors[train_idx, , drop = FALSE], stats::median, numeric(1), na.rm = TRUE)
    train_scale <- vapply(transformed_predictors[train_idx, , drop = FALSE], stats::IQR, numeric(1), na.rm = TRUE)
    train_sd <- vapply(transformed_predictors[train_idx, , drop = FALSE], stats::sd, numeric(1), na.rm = TRUE)
    train_scale[!is.finite(train_scale) | train_scale == 0] <- train_sd[!is.finite(train_scale) | train_scale == 0]
    train_scale[!is.finite(train_scale) | train_scale == 0] <- 1
    test_x <- as_tibble(sweep(sweep(as.matrix(test_x_raw), 2, train_center, "-"), 2, train_scale, "/"))

    pca_fold <- stats::prcomp(as.matrix(train_x), center = FALSE, scale. = FALSE)
    n_pcs <- min(8L, ncol(pca_fold$x), max(1L, floor(length(train_idx) / 10L)))
    train_scores <- as_tibble(pca_fold$x[, seq_len(n_pcs), drop = FALSE])
    test_scores <- as_tibble(as.matrix(test_x) %*% pca_fold$rotation[, seq_len(n_pcs), drop = FALSE])
    names(train_scores) <- paste0("PC", seq_len(n_pcs))
    names(test_scores) <- names(train_scores)
    train_scores$model_target <- target_log[train_idx]
    fold_fit <- stats::lm(model_target ~ ., data = train_scores)
    cv_pred_log[test_idx] <- as.numeric(stats::predict(fold_fit, newdata = test_scores))
    cv_n_pcs[test_idx] <- n_pcs
  }

  full_x <- transformed_predictors %>% robust_scale()
  full_pca <- stats::prcomp(as.matrix(full_x), center = FALSE, scale. = FALSE)
  full_n_pcs <- min(8L, ncol(full_pca$x), max(1L, floor(nrow(pip_keff_model_df) / 10L)))
  full_scores <- as_tibble(full_pca$x[, seq_len(full_n_pcs), drop = FALSE])
  names(full_scores) <- paste0("PC", seq_len(full_n_pcs))
  full_scores$model_target <- target_log
  full_fit <- stats::lm(model_target ~ ., data = full_scores)
  in_sample_pred_log <- as.numeric(stats::predict(full_fit, newdata = full_scores))

  pip_keff_prediction <- pip_keff_model_df %>%
    dplyr::mutate(
      predicted_log_pip_k_eff_cv = cv_pred_log,
      predicted_pip_k_eff_cv = pmax(0, expm1(.data$predicted_log_pip_k_eff_cv)),
      log1p_predicted_pip_k_eff_cv = log1p(.data$predicted_pip_k_eff_cv),
      log1p_pip_k_eff = log1p(.data$pip_k_eff),
      predicted_log_pip_k_eff_in_sample = in_sample_pred_log,
      predicted_pip_k_eff_in_sample = pmax(0, expm1(.data$predicted_log_pip_k_eff_in_sample)),
      pip_k_eff_residual = .data$pip_k_eff - .data$predicted_pip_k_eff_cv,
      log_pip_k_eff_residual = log1p(.data$pip_k_eff) - .data$predicted_log_pip_k_eff_cv,
      predicted_pip_k_eff_bin = dplyr::ntile(.data$predicted_pip_k_eff_cv, 3),
      actual_pip_k_eff_bin = dplyr::ntile(.data$pip_k_eff, 3),
      residual_bin = dplyr::ntile(.data$pip_k_eff_residual, 5),
      predicted_actual_zone = paste0(
        "pred", .data$predicted_pip_k_eff_bin,
        "_actual", .data$actual_pip_k_eff_bin
      ),
      residual_zone = dplyr::case_when(
        .data$residual_bin == 5L ~ "more_diffuse_than_predicted",
        .data$residual_bin == 1L ~ "less_diffuse_than_predicted",
        TRUE ~ "near_predicted"
      ),
      cv_n_pcs = cv_n_pcs
    )

  prediction_summary <- tibble(
    n_loci = nrow(pip_keff_prediction),
    n_predictors = length(pip_keff_predictor_cols),
    full_model_pcs = full_n_pcs,
    cv_spearman_actual_predicted = safe_cor(pip_keff_prediction$pip_k_eff, pip_keff_prediction$predicted_pip_k_eff_cv),
    cv_spearman_log_actual_predicted = safe_cor(log1p(pip_keff_prediction$pip_k_eff), pip_keff_prediction$predicted_log_pip_k_eff_cv),
    cv_rmse_log = sqrt(mean(pip_keff_prediction$log_pip_k_eff_residual^2, na.rm = TRUE)),
    cv_median_abs_residual = stats::median(abs(pip_keff_prediction$pip_k_eff_residual), na.rm = TRUE)
  )

  prediction_loadings <- as_tibble(
    full_pca$rotation[, seq_len(full_n_pcs), drop = FALSE],
    rownames = "metric"
  )

  residual_candidates <- pip_keff_prediction %>%
    dplyr::arrange(dplyr::desc(abs(.data$log_pip_k_eff_residual))) %>%
    dplyr::select(dplyr::all_of(c(
      "locus_id", "gene_name", "signal_pip_zone", "residual_zone",
      "max_pip", "pip_k_eff", "predicted_pip_k_eff_cv",
      "pip_k_eff_residual", "log_pip_k_eff_residual",
      "M1", "M2", "M4", "z_eff_signals_for_sampling"
    ))) %>%
    dplyr::slice_head(n = 75)

  pip_keff_prediction <- assign_interpretive_region(pip_keff_prediction)
  geometric_sample <- repulsive_geometric_sample(
    pip_keff_prediction,
    x_col = "log1p_predicted_pip_k_eff_cv",
    y_col = "log1p_pip_k_eff",
    n = GEOMETRIC_SAMPLE_N,
    forced_loci = GEOMETRIC_FORCED_LOCI
  ) %>%
    dplyr::select(dplyr::all_of(c(
      "geometric_sample_order", "locus_id", "gene_name",
      "suggested_sampling_region", "residual_zone", "signal_pip_zone",
      "max_pip", "pip_k_eff", "predicted_pip_k_eff_cv",
      "pip_k_eff_residual", "log_pip_k_eff_residual",
      "log1p_predicted_pip_k_eff_cv", "log1p_pip_k_eff",
      "M1", "M2", "M4", "z_eff_signals_for_sampling"
    )))

  residual_geometry_sample <- repulsive_geometric_sample(
    pip_keff_prediction,
    x_col = "log1p_predicted_pip_k_eff_cv",
    y_col = "log_pip_k_eff_residual",
    n = GEOMETRIC_SAMPLE_N,
    forced_loci = GEOMETRIC_FORCED_LOCI
  ) %>%
    dplyr::rename(residual_geometry_sample_order = geometric_sample_order) %>%
    dplyr::select(dplyr::all_of(c(
      "residual_geometry_sample_order", "locus_id", "gene_name",
      "suggested_sampling_region", "residual_zone", "signal_pip_zone",
      "max_pip", "pip_k_eff", "predicted_pip_k_eff_cv",
      "pip_k_eff_residual", "log_pip_k_eff_residual",
      "log1p_predicted_pip_k_eff_cv", "log1p_pip_k_eff",
      "M1", "M2", "M4", "z_eff_signals_for_sampling"
    )))

  pip_keff_prediction <- pip_keff_prediction %>%
    dplyr::left_join(
      geometric_sample %>%
        dplyr::select(dplyr::all_of(c("locus_id", "geometric_sample_order"))),
      by = "locus_id"
    ) %>%
    dplyr::left_join(
      residual_geometry_sample %>%
        dplyr::select(dplyr::all_of(c("locus_id", "residual_geometry_sample_order"))),
      by = "locus_id"
    ) %>%
    dplyr::mutate(
      geometric_sample_selected = !is.na(.data$geometric_sample_order),
      residual_geometry_sample_selected = !is.na(.data$residual_geometry_sample_order)
    )

  soft_suggested_counts <- tibble::tribble(
    ~suggested_sampling_region, ~soft_suggested_n,
    "expected_diffuse", 4L,
    "surprisingly_diffuse", 4L,
    "surprisingly_concentrated", 4L,
    "typical_or_transition", 4L,
    "easy_control", 2L,
    "low_predicted_diffuse_outlier", 2L
  )

  geometric_sample_counts <- pip_keff_prediction %>%
    dplyr::count(.data$suggested_sampling_region, name = "available_n") %>%
    dplyr::full_join(
      pip_keff_prediction %>%
        dplyr::filter(.data$geometric_sample_selected) %>%
        dplyr::count(.data$suggested_sampling_region, name = "geometric_selected_n"),
      by = "suggested_sampling_region"
    ) %>%
    dplyr::full_join(soft_suggested_counts, by = "suggested_sampling_region") %>%
    dplyr::mutate(
      available_n = dplyr::coalesce(.data$available_n, 0L),
      geometric_selected_n = dplyr::coalesce(.data$geometric_selected_n, 0L),
      soft_suggested_n = dplyr::coalesce(.data$soft_suggested_n, 0L),
      forced_anchor_selected_n = sum(geometric_sample$locus_id %in% GEOMETRIC_FORCED_LOCI)
    ) %>%
    dplyr::arrange(dplyr::desc(.data$soft_suggested_n), .data$suggested_sampling_region)

  residual_geometry_sample_counts <- pip_keff_prediction %>%
    dplyr::count(.data$suggested_sampling_region, name = "available_n") %>%
    dplyr::full_join(
      pip_keff_prediction %>%
        dplyr::filter(.data$residual_geometry_sample_selected) %>%
        dplyr::count(.data$suggested_sampling_region, name = "residual_geometry_selected_n"),
      by = "suggested_sampling_region"
    ) %>%
    dplyr::full_join(soft_suggested_counts, by = "suggested_sampling_region") %>%
    dplyr::mutate(
      available_n = dplyr::coalesce(.data$available_n, 0L),
      residual_geometry_selected_n = dplyr::coalesce(.data$residual_geometry_selected_n, 0L),
      soft_suggested_n = dplyr::coalesce(.data$soft_suggested_n, 0L),
      forced_anchor_selected_n = sum(residual_geometry_sample$locus_id %in% GEOMETRIC_FORCED_LOCI)
    ) %>%
    dplyr::arrange(dplyr::desc(.data$soft_suggested_n), .data$suggested_sampling_region)

  geometric_sample_comparison_counts <- geometric_sample_counts %>%
    dplyr::select(dplyr::all_of(c(
      "suggested_sampling_region", "available_n",
      "geometric_selected_n", "soft_suggested_n"
    ))) %>%
    dplyr::full_join(
      residual_geometry_sample_counts %>%
        dplyr::select(dplyr::all_of(c(
          "suggested_sampling_region", "residual_geometry_selected_n"
        ))),
      by = "suggested_sampling_region"
    ) %>%
    dplyr::mutate(
      geometric_selected_n = dplyr::coalesce(.data$geometric_selected_n, 0L),
      residual_geometry_selected_n = dplyr::coalesce(.data$residual_geometry_selected_n, 0L)
    ) %>%
    dplyr::arrange(dplyr::desc(.data$soft_suggested_n), .data$suggested_sampling_region)

  readr::write_csv(pip_keff_prediction, PIP_KEFF_PREDICTION_PATH)
  readr::write_csv(prediction_loadings, PIP_KEFF_PREDICTION_LOADINGS_PATH)
  readr::write_csv(prediction_summary, PIP_KEFF_PREDICTION_SUMMARY_PATH)
  readr::write_csv(residual_candidates, PIP_KEFF_RESIDUAL_CANDIDATES_PATH)
  readr::write_csv(geometric_sample, PIP_KEFF_GEOMETRIC_SAMPLE_PATH)
  readr::write_csv(geometric_sample_counts, PIP_KEFF_GEOMETRIC_SAMPLE_COUNTS_PATH)
  readr::write_csv(residual_geometry_sample, PIP_KEFF_RESIDUAL_GEOMETRIC_SAMPLE_PATH)
  readr::write_csv(residual_geometry_sample_counts, PIP_KEFF_RESIDUAL_GEOMETRIC_SAMPLE_COUNTS_PATH)
  readr::write_csv(geometric_sample_comparison_counts, PIP_KEFF_GEOMETRIC_SAMPLE_COMPARISON_COUNTS_PATH)

  locked_geometric_n20 <- geometric_sample %>%
    dplyr::arrange(.data$geometric_sample_order)
  if (nrow(locked_geometric_n20) != GEOMETRIC_SAMPLE_N ||
      dplyr::n_distinct(locked_geometric_n20$locus_id) != GEOMETRIC_SAMPLE_N ||
      !identical(as.integer(locked_geometric_n20$geometric_sample_order), seq_len(GEOMETRIC_SAMPLE_N))) {
    stop("Locked geometric sample must contain exactly ", GEOMETRIC_SAMPLE_N,
         " unique loci ordered 1..", GEOMETRIC_SAMPLE_N, ".")
  }
  if (!file.exists(MANIFEST_PATH)) {
    stop("Manifest not found for locked sample export: ", MANIFEST_PATH)
  }
  manifest_for_lock <- readr::read_csv(MANIFEST_PATH, show_col_types = FALSE)
  locked_manifest <- locked_geometric_n20 %>%
    dplyr::left_join(manifest_for_lock, by = "locus_id", suffix = c("_sample", ""))
  missing_manifest_ids <- locked_manifest %>%
    dplyr::filter(is.na(.data$gene_id)) %>%
    dplyr::pull(.data$locus_id)
  if (length(missing_manifest_ids) > 0) {
    stop("Locked sample contains loci absent from manifest: ",
         paste(missing_manifest_ids, collapse = ", "))
  }
  annotation_selection_geometric_n20 <- locked_geometric_n20 %>%
    dplyr::transmute(
      locus_id = .data$locus_id,
      annotate = TRUE,
      notes = "baseline_susie_predicted_actual_geometric_n20",
      priority = as.integer(.data$geometric_sample_order),
      annotation_gene_name_override = NA_character_,
      annotation_gene_id_override = NA_character_,
      annotation_tissue_override = NA_character_
    )
  readr::write_csv(locked_geometric_n20, LOCKED_GEOMETRIC_N20_PATH)
  readr::write_csv(locked_manifest, LOCKED_GEOMETRIC_N20_MANIFEST_PATH)
  readr::write_csv(annotation_selection_geometric_n20, LOCKED_GEOMETRIC_N20_SELECTION_PATH)
  readr::write_csv(annotation_selection_geometric_n20, CONFIG_GEOMETRIC_N20_SELECTION_PATH)

  label_df <- pip_keff_prediction %>%
    dplyr::filter(.data$locus_id %in% LABEL_LOCI | .data$locus_id %in% head(residual_candidates$locus_id, 8L))
  geometric_label_df <- pip_keff_prediction %>%
    dplyr::filter(.data$geometric_sample_selected)
  residual_geometry_label_df <- pip_keff_prediction %>%
    dplyr::filter(.data$residual_geometry_sample_selected)

  p_pred_actual <- ggplot(
    pip_keff_prediction,
    aes(x = .data$log1p_predicted_pip_k_eff_cv, y = .data$log1p_pip_k_eff, color = .data$residual_zone)
  ) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey45") +
    geom_point(aes(size = .data$max_pip), alpha = 0.78) +
    geom_point(
      data = pip_keff_prediction %>% dplyr::filter(.data$geometric_sample_selected),
      shape = 21,
      fill = "#FFD166",
      color = "black",
      size = 3.2,
      stroke = 0.75,
      inherit.aes = FALSE,
      aes(x = .data$log1p_predicted_pip_k_eff_cv, y = .data$log1p_pip_k_eff)
    ) +
    geom_text(
      data = label_df,
      aes(label = .data$locus_id),
      size = 3,
      nudge_y = 0.04,
      check_overlap = TRUE
    ) +
    labs(
      title = "Predicted vs actual baseline SuSiE PIP effective size",
      subtitle = sprintf(
        "5-fold CV; Spearman rho = %.2f; positive residual = more diffuse than expected",
        prediction_summary$cv_spearman_actual_predicted[[1]]
      ),
      x = "log1p(predicted pip_k_eff from dataset metrics)",
      y = "log1p(actual baseline SuSiE pip_k_eff)",
      color = "Residual zone",
      size = "max PIP"
    ) +
    theme(legend.position = "bottom")

  p_geometric_sample <- ggplot(
    pip_keff_prediction,
    aes(x = .data$log1p_predicted_pip_k_eff_cv, y = .data$log1p_pip_k_eff)
  ) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey45") +
    geom_point(color = "grey78", alpha = 0.55, size = 1.8) +
    geom_point(
      data = pip_keff_prediction %>% dplyr::filter(.data$geometric_sample_selected),
      aes(fill = .data$suggested_sampling_region, size = .data$max_pip),
      shape = 21,
      color = "black",
      stroke = 0.9,
      alpha = 0.95
    ) +
    geom_text(
      data = geometric_label_df,
      aes(label = .data$geometric_sample_order),
      size = 2.8,
      color = "black"
    ) +
    labs(
      title = sprintf("Geometric maximin-repulsion sample of %d loci", GEOMETRIC_SAMPLE_N),
      subtitle = "Selection uses only the plotted predicted-vs-actual coordinates; region labels are diagnostic only.",
      x = "log1p(predicted pip_k_eff from dataset metrics)",
      y = "log1p(actual baseline SuSiE pip_k_eff)",
      fill = "Diagnostic region",
      size = "max PIP"
    ) +
    theme(legend.position = "bottom")

  p_residual_geometry_sample <- ggplot(
    pip_keff_prediction,
    aes(x = .data$log1p_predicted_pip_k_eff_cv, y = .data$log1p_pip_k_eff)
  ) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey45") +
    geom_point(color = "grey78", alpha = 0.55, size = 1.8) +
    geom_point(
      data = pip_keff_prediction %>% dplyr::filter(.data$residual_geometry_sample_selected),
      aes(fill = .data$suggested_sampling_region, size = .data$max_pip),
      shape = 21,
      color = "black",
      stroke = 0.9,
      alpha = 0.95
    ) +
    geom_text(
      data = residual_geometry_label_df,
      aes(label = .data$residual_geometry_sample_order),
      size = 2.8,
      color = "black"
    ) +
    labs(
      title = sprintf("Residual-geometry maximin sample of %d loci", GEOMETRIC_SAMPLE_N),
      subtitle = "Selection uses predicted diffuseness and residual diffuseness; plotted back on predicted vs actual axes.",
      x = "log1p(predicted pip_k_eff from dataset metrics)",
      y = "log1p(actual baseline SuSiE pip_k_eff)",
      fill = "Diagnostic region",
      size = "max PIP"
    ) +
    theme(legend.position = "bottom")

  p_residual <- ggplot(
    pip_keff_prediction,
    aes(x = .data$log1p_predicted_pip_k_eff_cv, y = .data$log_pip_k_eff_residual, color = .data$signal_pip_zone)
  ) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "grey45") +
    geom_point(aes(size = .data$max_pip), alpha = 0.78) +
    geom_text(
      data = label_df,
      aes(label = .data$locus_id),
      size = 3,
      nudge_y = 0.04,
      check_overlap = TRUE
    ) +
    labs(
      title = "PIP effective-size residuals vs predicted diffuseness",
      x = "log1p(predicted pip_k_eff from dataset metrics)",
      y = "log1p(actual pip_k_eff) - log1p(predicted pip_k_eff)",
      color = "Signal/PIP zone",
      size = "max PIP"
    ) +
    theme(legend.position = "bottom")

  p_residual_hist <- ggplot(pip_keff_prediction, aes(x = .data$log_pip_k_eff_residual, fill = .data$residual_zone)) +
    geom_histogram(bins = 35, color = "white", linewidth = 0.2) +
    labs(
      title = "Prediction residual distribution for PIP effective size",
      x = "log residual",
      y = "Locus count",
      fill = "Residual zone"
    ) +
    theme(legend.position = "bottom")

  ggsave(file.path(FIGURE_DIR, "pip_k_eff_predicted_vs_actual.png"), p_pred_actual, width = 8, height = 6, dpi = 150)
  ggsave(file.path(FIGURE_DIR, "pip_k_eff_geometric_sample_n20.png"), p_geometric_sample, width = 9, height = 7, dpi = 150)
  ggsave(file.path(FIGURE_DIR, "pip_k_eff_residual_geometry_sample_n20.png"), p_residual_geometry_sample, width = 9, height = 7, dpi = 150)
  ggsave(file.path(FIGURE_DIR, "pip_k_eff_residual_vs_predicted.png"), p_residual, width = 8, height = 6, dpi = 150)
  ggsave(file.path(FIGURE_DIR, "pip_k_eff_prediction_residual_histogram.png"), p_residual_hist, width = 8, height = 6, dpi = 150)

  list(
    prediction_summary = prediction_summary,
    predictor_columns = pip_keff_predictor_cols,
    geometric_sample_comparison_counts = geometric_sample_comparison_counts,
    geometric_sample_counts = geometric_sample_counts,
    geometric_sample_preview = geometric_sample %>% dplyr::slice_head(n = 20),
    residual_geometry_sample_counts = residual_geometry_sample_counts,
    residual_geometry_sample_preview = residual_geometry_sample %>% dplyr::slice_head(n = 20),
    residual_candidate_preview = residual_candidates %>% dplyr::slice_head(n = 15)
  )
  p_pred_actual
  p_geometric_sample
  p_residual_geometry_sample
  p_residual
  p_residual_hist
} else {
  message("Skipping pip_k_eff prediction: need at least 20 complete loci and 2 usable predictors.")
}
```

# 9. Zone Diagnostics

```{r zone-plots, fig.width=14, fig.height=4.5}
plot_df <- joined_metrics %>%
  dplyr::filter(.data$qc_status == "qc_pass", .data$status == "completed")

make_zone_plot <- function(x_col, y_col, title) {
  if (!all(c(x_col, y_col) %in% names(plot_df))) {
    return(
      ggplot() +
        annotate("text", x = 0, y = 0, label = paste("Missing columns for", title)) +
        theme_void()
    )
  }
  p <- ggplot(plot_df, aes(x = .data[[x_col]], y = .data[[y_col]], color = .data$signal_pip_zone)) +
    geom_point(alpha = 0.8, size = 2) +
    label_loci_layer(plot_df, x_col, y_col) +
    labs(title = title, x = x_col, y = y_col, color = "Zone") +
    theme(legend.position = "bottom")
  if (identical(y_col, "max_pip")) {
    p <- p + geom_hline(yintercept = PIP_HIGH_THRESHOLD, linetype = "dashed", color = "grey45")
  }
  if (identical(x_col, "z_max_abs")) {
    p <- p + geom_vline(xintercept = Z_HIGH_THRESHOLD, linetype = "dashed", color = "grey45")
  }
  p
}

p_zone_1 <- make_zone_plot("z_max_abs", "max_pip", "Marginal signal vs max PIP")
p_zone_2 <- make_zone_plot("z_eff_signals_for_sampling", "pip_k_eff", "z effective signals vs PIP effective size")
p_zone_3 <- make_zone_plot("M1", "max_pip", "LD mid-energy vs max PIP")

p_zone_1
p_zone_2
p_zone_3

ggsave(file.path(FIGURE_DIR, "zone_z_max_abs_vs_max_pip.png"), p_zone_1, width = 8, height = 6, dpi = 150)
ggsave(file.path(FIGURE_DIR, "zone_z_eff_signals_vs_pip_k_eff.png"), p_zone_2, width = 8, height = 6, dpi = 150)
ggsave(file.path(FIGURE_DIR, "zone_M1_vs_max_pip.png"), p_zone_3, width = 8, height = 6, dpi = 150)
```

# 10. Loci of Interest

```{r loci-of-interest}
joined_metrics %>%
  dplyr::filter(.data$locus_id %in% LABEL_LOCI) %>%
  dplyr::select(dplyr::any_of(c(
    "locus_id", "gene_name", "qc_status", "status",
    "M1", "M2", "M4",
    "z_max_abs", "z_topk_ratio", "z_eff_signals",
    "max_pip", "n_pip_gt_0_5", "n_pip_gt_0_9",
    "pip_k_eff", "pip_top2_gap",
    "n_active_effects",
    "median_active_alpha_k_eff_core95",
    "signal_pip_zone"
  )))
```
