---
title: "5. Run Vanilla SuSiE RSS on Selected Loci"
output: html_notebook
---

```{r setup}
required_packages <- c("arrow", "dplyr", "ggplot2", "patchwork", "readr", "susieR", "tibble", "tidyr")
missing_packages <- required_packages[!vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing_packages) > 0) {
  stop(
    "Missing required R packages: ",
    paste(missing_packages, collapse = ", "),
    ". Install them before running this notebook."
  )
}

library(arrow)
library(dplyr)
library(ggplot2)
library(patchwork)
library(readr)
library(susieR)
library(tibble)
library(tidyr)

find_project_root <- function(start = getwd()) {
  current <- normalizePath(start, winslash = "/", mustWork = TRUE)
  # Last entry is an example fallback path -- set to your environment.
  candidates <- unique(c(current, dirname(current), dirname(dirname(current)),
                         "/path/to/eQTL_annotations_for_susine"))

  for (candidate in candidates) {
    if (
      dir.exists(file.path(candidate, "utils")) &&
      dir.exists(file.path(candidate, "vignettes")) &&
      file.exists(file.path(candidate, "README.md"))
    ) {
      return(candidate)
    }
  }

  stop("Could not locate eQTL_annotations_for_susine project root")
}

resolve_selection_path <- function(project_root) {
  selection_dir <- file.path(project_root, "output", "prelim", "phase1_metrics_screening_review")
  candidates <- sort(list.files(
    selection_dir,
    pattern = "representative_gene_sample_n.*_annotation_selection\\.csv$",
    full.names = TRUE
  ))
  if (length(candidates) == 0) {
    stop("Could not find a selected-loci annotation selection CSV under ", selection_dir)
  }
  candidates[[length(candidates)]]
}

merge_manifest_selection <- function(manifest_df, selection_df) {
  if (!"locus_id" %in% names(manifest_df) || !"locus_id" %in% names(selection_df)) {
    stop("Both manifest and selection files must include a locus_id column.")
  }

  merged_df <- manifest_df %>%
    inner_join(selection_df, by = "locus_id", suffix = c("_manifest", "_selection"))

  for (field in c("gene_name", "gene_id", "gtex_tissue", "gtex_chrom", "priority", "notes")) {
    manifest_col <- paste0(field, "_manifest")
    selection_col <- paste0(field, "_selection")
    if (manifest_col %in% names(merged_df) && selection_col %in% names(merged_df)) {
      merged_df[[field]] <- dplyr::coalesce(merged_df[[selection_col]], merged_df[[manifest_col]])
    } else if (selection_col %in% names(merged_df)) {
      merged_df[[field]] <- merged_df[[selection_col]]
    } else if (manifest_col %in% names(merged_df)) {
      merged_df[[field]] <- merged_df[[manifest_col]]
    }
  }

  merged_df
}

build_dense_ld <- function(ld_long_df, p) {
  R <- matrix(0, nrow = p, ncol = p)
  diag(R) <- 1
  idx1 <- ld_long_df$snp_index_1 + 1L
  idx2 <- ld_long_df$snp_index_2 + 1L
  R[cbind(idx1, idx2)] <- ld_long_df$r
  R[cbind(idx2, idx1)] <- ld_long_df$r
  R
}

compute_posterior_bhat <- function(susie_fit) {
  if (is.null(susie_fit$alpha) || is.null(susie_fit$mu)) {
    stop("susie_fit is missing alpha or mu; could not compute posterior b_hat.")
  }
  colSums(susie_fit$alpha * susie_fit$mu)
}

PROJECT_ROOT <- find_project_root()
MANIFEST_PATH <- file.path(PROJECT_ROOT, "config", "loci_manifest_sample_100_per_chrom.csv")
SELECTION_PATH <- resolve_selection_path(PROJECT_ROOT)
OUTPUT_DIR <- file.path(PROJECT_ROOT, "output")
LD_DIR <- file.path(OUTPUT_DIR, "ld")
REVIEW_DIR <- file.path(OUTPUT_DIR, "susie_rss", tools::file_path_sans_ext(basename(SELECTION_PATH)))
dir.create(REVIEW_DIR, recursive = TRUE, showWarnings = FALSE)

SUSIE_L <- 10
ESTIMATE_RESIDUAL_VARIANCE <- TRUE
CHECK_PRIOR <- FALSE

cat("Project root:", PROJECT_ROOT, "\n")
cat("Manifest:", MANIFEST_PATH, "\n")
cat("Selection:", SELECTION_PATH, "\n")
cat("Review dir:", REVIEW_DIR, "\n")
```

## Scope

This notebook runs vanilla `susieR::susie_rss()` on every selected locus using the stored Phase 1:

- `z_score` values from each locus master variant file
- LD order TSV
- long-form LD parquet

No additional data prep is needed as long as those three artifacts exist and the LD matrix reconstructs cleanly.

```{r load-selection}
manifest_df <- readr::read_csv(MANIFEST_PATH, show_col_types = FALSE)
selection_df <- readr::read_csv(SELECTION_PATH, show_col_types = FALSE)
selected_loci_df <- merge_manifest_selection(manifest_df, selection_df) %>%
  distinct(locus_id, .keep_all = TRUE)

if ("priority" %in% names(selected_loci_df)) {
  selected_loci_df <- selected_loci_df %>% arrange(priority, locus_id)
} else {
  selected_loci_df <- selected_loci_df %>% arrange(locus_id)
}

selected_loci_df %>%
  select(any_of(c("locus_id", "gene_name", "gtex_tissue", "gtex_chrom", "priority", "notes")))
```

```{r run-susie}
run_rows <- list()
bhat_rows <- list()
top_rows <- list()
error_rows <- list()
run_summary_cols <- c(
  "locus_id", "gene_name", "n_variants", "n_rss", "elbo_last",
  "cs_count", "max_pip", "mean_abs_bhat", "max_abs_bhat",
  "status", "error_message"
)

for (i in seq_len(nrow(selected_loci_df))) {
  locus_row <- selected_loci_df[i, ]
  locus_id <- locus_row$locus_id[[1]]
  gene_name <- locus_row$gene_name[[1]]

  locus_dir <- file.path(LD_DIR, locus_id)
  master_path <- file.path(locus_dir, sprintf("%s_phase1_master_variants.csv", gene_name))
  order_path <- file.path(locus_dir, sprintf("%s_LD_variant_order.tsv", gene_name))
  ld_long_path <- file.path(locus_dir, sprintf("%s_phase1_LD_R_long.parquet", gene_name))

  cat(sprintf("[%d/%d] %s (%s)\n", i, nrow(selected_loci_df), locus_id, gene_name))

  if (!file.exists(master_path) || !file.exists(order_path) || !file.exists(ld_long_path)) {
    error_rows[[length(error_rows) + 1L]] <- tibble(
      locus_id = locus_id,
      gene_name = gene_name,
      status = "missing_inputs",
      error_message = paste(
        c(
          if (!file.exists(master_path)) master_path,
          if (!file.exists(order_path)) order_path,
          if (!file.exists(ld_long_path)) ld_long_path
        ),
        collapse = " | "
      )
    )
    next
  }

  tryCatch({
    master_df <- readr::read_csv(master_path, show_col_types = FALSE) %>%
      filter(.data$ld_included) %>%
      transmute(
        variant_id,
        z_score,
        slope,
        slope_se,
        sample_size,
        ld_matrix_index
      )

    order_df <- readr::read_tsv(order_path, show_col_types = FALSE)
    ld_long_df <- arrow::read_parquet(ld_long_path) %>% as_tibble()

    aligned_df <- order_df %>%
      left_join(master_df, by = c("id" = "variant_id")) %>%
      arrange(.data$index)

    if (any(is.na(aligned_df$z_score))) {
      stop("Aligned z_score vector contains NA values.")
    }

    expected_index <- seq_len(nrow(aligned_df)) - 1L
    if (!all(aligned_df$index == expected_index)) {
      stop("LD order index is not contiguous from 0 to p-1.")
    }

    p <- nrow(aligned_df)
    R <- build_dense_ld(ld_long_df, p)

    if (max(abs(R - t(R))) > 1e-8) {
      stop("Reconstructed LD matrix is not symmetric within tolerance.")
    }
    if (max(abs(diag(R) - 1)) > 1e-8) {
      stop("Reconstructed LD matrix diagonal is not all ones within tolerance.")
    }

    n_rss <- round(mean(aligned_df$sample_size, na.rm = TRUE))
    susie_fit <- susieR::susie_rss(
      z = aligned_df$z_score,
      R = R,
      n = n_rss,
      L = SUSIE_L,
      estimate_residual_variance = ESTIMATE_RESIDUAL_VARIANCE,
      check_prior = CHECK_PRIOR
    )

    posterior_bhat <- compute_posterior_bhat(susie_fit)

    locus_bhat_df <- aligned_df %>%
      transmute(
        locus_id = locus_id,
        gene_name = gene_name,
        variant_id = id,
        z_score,
        slope,
        slope_se,
        sample_size,
        posterior_bhat = posterior_bhat,
        pip = susie_fit$pip
      )

    top10_df <- locus_bhat_df %>%
      arrange(desc(abs(.data$posterior_bhat))) %>%
      mutate(rank_abs_bhat = row_number()) %>%
      slice_head(n = 10)

    run_rows[[length(run_rows) + 1L]] <- tibble(
      locus_id = locus_id,
      gene_name = gene_name,
      n_variants = p,
      n_rss = n_rss,
      elbo_last = tail(susie_fit$elbo, 1),
      cs_count = length(susie_fit$sets$cs),
      max_pip = max(susie_fit$pip, na.rm = TRUE),
      mean_abs_bhat = mean(abs(posterior_bhat), na.rm = TRUE),
      max_abs_bhat = max(abs(posterior_bhat), na.rm = TRUE),
      status = "completed",
      error_message = NA_character_
    )

    bhat_rows[[length(bhat_rows) + 1L]] <- locus_bhat_df
    top_rows[[length(top_rows) + 1L]] <- top10_df
  }, error = function(e) {
    error_rows[[length(error_rows) + 1L]] <<- tibble(
      locus_id = locus_id,
      gene_name = gene_name,
      status = "failed",
      error_message = conditionMessage(e)
    )
  })
}

susie_run_summary_df <- bind_rows(run_rows)
susie_errors_df <- bind_rows(error_rows)
posterior_bhat_df <- bind_rows(bhat_rows)
top10_bhat_df <- bind_rows(top_rows)

if (nrow(susie_run_summary_df) == 0) {
  susie_run_summary_df <- tibble(
    locus_id = character(),
    gene_name = character(),
    n_variants = integer(),
    n_rss = integer(),
    elbo_last = double(),
    cs_count = integer(),
    max_pip = double(),
    mean_abs_bhat = double(),
    max_abs_bhat = double(),
    status = character(),
    error_message = character()
  )
}

if (nrow(susie_errors_df) > 0) {
  susie_run_summary_df <- bind_rows(
    susie_run_summary_df,
    susie_errors_df %>% mutate(
      n_variants = NA_integer_,
      n_rss = NA_integer_,
      elbo_last = NA_real_,
      cs_count = NA_integer_,
      max_pip = NA_real_,
      mean_abs_bhat = NA_real_,
      max_abs_bhat = NA_real_
    ) %>% select(all_of(run_summary_cols))
  )
}

write_csv(susie_run_summary_df, file.path(REVIEW_DIR, "susie_rss_run_summary.csv"))
write_csv(posterior_bhat_df, file.path(REVIEW_DIR, "susie_rss_posterior_bhat_table.csv"))
write_csv(top10_bhat_df, file.path(REVIEW_DIR, "susie_rss_top10_posterior_bhat_per_locus.csv"))

susie_run_summary_df
```

## Failed loci

```{r failed-loci}
susie_errors_df
```

## Posterior `b_hat` distribution across loci

```{r bhat-across-loci, fig.width=12, fig.height=8}
posterior_bhat_df %>%
  ggplot(aes(x = posterior_bhat)) +
  geom_histogram(bins = 40, fill = "#2f6c8f", color = "white") +
  facet_wrap(~ locus_id, scales = "free_y") +
  labs(
    title = "Posterior b_hat distribution across selected loci",
    x = "Posterior b_hat = colSums(alpha * mu)",
    y = "Variant count"
  ) +
  theme_minimal(base_size = 12)
```

## Top 10 absolute `b_hat` variants per locus

```{r top10-table}
top10_bhat_df %>%
  arrange(locus_id, rank_abs_bhat) %>%
  select(locus_id, gene_name, rank_abs_bhat, variant_id, z_score, slope, slope_se, posterior_bhat, pip)
```

## Per-locus histograms with top 10 markers

```{r per-locus-hists, fig.width=10, fig.height=6}
plot_manifest <- list()

for (locus_id in unique(posterior_bhat_df$locus_id)) {
  locus_bhat_df <- posterior_bhat_df %>% filter(.data$locus_id == !!locus_id)
  locus_top10_df <- top10_bhat_df %>% filter(.data$locus_id == !!locus_id)
  if (nrow(locus_bhat_df) == 0) {
    next
  }

  gene_name <- unique(locus_bhat_df$gene_name)[[1]]

  p_hist <- ggplot(locus_bhat_df, aes(x = posterior_bhat)) +
    geom_histogram(bins = 40, fill = "#2f6c8f", color = "white") +
    geom_vline(
      data = locus_top10_df,
      aes(xintercept = posterior_bhat),
      color = "#b85c38",
      linetype = "dashed",
      linewidth = 0.7,
      alpha = 0.8
    ) +
    labs(
      title = sprintf("%s (%s): posterior b_hat distribution", gene_name, locus_id),
      x = "Posterior b_hat = colSums(alpha * mu)",
      y = "Variant count"
    ) +
    theme_minimal(base_size = 12)

  # print(p_hist)

  plot_path <- file.path(REVIEW_DIR, sprintf("%s_%s_posterior_bhat_histogram.png", gene_name, locus_id))
  ggsave(plot_path, p_hist, width = 10, height = 6, dpi = 150)
  plot_manifest[[length(plot_manifest) + 1L]] <- tibble(
    locus_id = locus_id,
    gene_name = gene_name,
    plot_path = plot_path
  )
}

plot_manifest_df <- bind_rows(plot_manifest)
write_csv(plot_manifest_df, file.path(REVIEW_DIR, "susie_rss_plot_manifest.csv"))
plot_manifest_df
```

## Notes

- This workbook runs plain `susie_rss` only. No AlphaGenome or other annotation vector is used here.
- Posterior `b_hat` is defined here as `colSums(alpha * mu)` from the `susie_rss` fit.
- If a locus fails here, the most likely causes are:
  - missing Phase 1 LD artifacts
  - NA values in `z_score`
  - malformed or non-symmetric reconstructed LD
