---
title: "6. Run Baseline SuSiNE RSS on Selected Loci"
output: html_notebook
---

```{r setup}
required_packages <- c(
  "arrow", "devtools", "dplyr", "ggplot2", "patchwork", "readr", "tibble", "tidyr", "purrr"
)
missing_packages <- required_packages[!vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing_packages) > 0) {
  stop(
    "Missing required R packages: ",
    paste(missing_packages, collapse = ", "),
    ". Install them before running this notebook."
  )
}

library(arrow)
library(devtools)
library(dplyr)
library(ggplot2)
library(patchwork)
library(readr)
library(tibble)
library(tidyr)
library(purrr)

find_project_root <- function(start = getwd()) {
  current <- normalizePath(start, winslash = "/", mustWork = TRUE)
  candidates <- unique(c(
    current,
    dirname(current),
    dirname(dirname(current)),
    # Example fallback path -- set to your environment.
    "/path/to/eQTL_annotations_for_susine"
  ))

  for (candidate in candidates) {
    if (
      dir.exists(file.path(candidate, "utils")) &&
      dir.exists(file.path(candidate, "vignettes")) &&
      file.exists(file.path(candidate, "README.md"))
    ) {
      return(candidate)
    }
  }

  stop("Could not locate eQTL_annotations_for_susine project root")
}

resolve_selection_path <- function(project_root) {
  selection_dir <- file.path(project_root, "output", "prelim", "phase1_metrics_screening_review")
  candidates <- sort(list.files(
    selection_dir,
    pattern = "representative_gene_sample_n.*_annotation_selection\\.csv$",
    full.names = TRUE
  ))
  if (length(candidates) == 0) {
    stop("Could not find a selected-loci annotation selection CSV under ", selection_dir)
  }
  candidates[[length(candidates)]]
}

merge_manifest_selection <- function(manifest_df, selection_df) {
  merged_df <- manifest_df %>%
    inner_join(selection_df, by = "locus_id", suffix = c("_manifest", "_selection"))

  for (field in c("gene_name", "gene_id", "gtex_tissue", "gtex_chrom", "priority", "notes")) {
    manifest_col <- paste0(field, "_manifest")
    selection_col <- paste0(field, "_selection")
    if (manifest_col %in% names(merged_df) && selection_col %in% names(merged_df)) {
      merged_df[[field]] <- dplyr::coalesce(merged_df[[selection_col]], merged_df[[manifest_col]])
    } else if (selection_col %in% names(merged_df)) {
      merged_df[[field]] <- merged_df[[selection_col]]
    } else if (manifest_col %in% names(merged_df)) {
      merged_df[[field]] <- merged_df[[manifest_col]]
    }
  }

  merged_df
}

build_dense_ld <- function(ld_long_df, p) {
  R <- matrix(0, nrow = p, ncol = p)
  diag(R) <- 1
  idx1 <- ld_long_df$snp_index_1 + 1L
  idx2 <- ld_long_df$snp_index_2 + 1L
  R[cbind(idx1, idx2)] <- ld_long_df$r
  R[cbind(idx2, idx1)] <- ld_long_df$r
  R
}

rms <- function(x) {
  x <- as.numeric(x)
  x <- x[is.finite(x)]
  if (!length(x)) return(NA_real_)
  sqrt(mean(x^2))
}

approx_mode <- function(x) {
  x <- as.numeric(x)
  x <- x[is.finite(x)]
  if (!length(x)) return(NA_real_)
  if (length(unique(x)) == 1) return(x[[1]])
  hist_obj <- hist(x, breaks = pretty(range(x), n = 40), plot = FALSE)
  max_bin <- which.max(hist_obj$counts)
  mean(hist_obj$breaks[c(max_bin, max_bin + 1)])
}

summarise_distribution <- function(x) {
  x <- as.numeric(x)
  x <- x[is.finite(x)]
  if (!length(x)) {
    return(tibble(
      mean = NA_real_,
      median = NA_real_,
      mode = NA_real_,
      q25 = NA_real_,
      q75 = NA_real_,
      q90 = NA_real_,
      q95 = NA_real_,
      variance = NA_real_
    ))
  }
  qs <- stats::quantile(x, probs = c(0.25, 0.75, 0.9, 0.95), na.rm = TRUE, names = FALSE)
  tibble(
    mean = mean(x),
    median = median(x),
    mode = approx_mode(x),
    q25 = qs[[1]],
    q75 = qs[[2]],
    q90 = qs[[3]],
    q95 = qs[[4]],
    variance = stats::var(x)
  )
}

extract_beta_hat_diagnostics <- function(fit, fit_log) {
  if (!is.null(fit$beta_hat_diagnostics) && nrow(as.data.frame(fit$beta_hat_diagnostics)) > 0) {
    return(as_tibble(fit$beta_hat_diagnostics))
  }
  if (!is.null(fit$model_fit$beta_hat_diagnostics) && nrow(as.data.frame(fit$model_fit$beta_hat_diagnostics)) > 0) {
    return(as_tibble(fit$model_fit$beta_hat_diagnostics))
  }

  beta_lines <- grep("^\\[beta_hat\\]", fit_log, value = TRUE)
  if (!length(beta_lines)) {
    return(tibble())
  }

  parse_line <- function(line) {
    pattern <- paste0(
      "^\\[beta_hat\\] iter=([0-9]+) effect=([0-9]+) ",
      "mean=([-0-9.eE]+) median=([-0-9.eE]+) mode=([-0-9.eE]+) ",
      "q25=([-0-9.eE]+) q75=([-0-9.eE]+) q90=([-0-9.eE]+) q95=([-0-9.eE]+) var=([-0-9.eE]+)$"
    )
    m <- regexec(pattern, line)
    parts <- regmatches(line, m)[[1]]
    if (length(parts) != 11) {
      return(NULL)
    }
    tibble(
      iteration = as.integer(parts[[2]]),
      effect = as.integer(parts[[3]]),
      mean = as.numeric(parts[[4]]),
      median = as.numeric(parts[[5]]),
      mode = as.numeric(parts[[6]]),
      q25 = as.numeric(parts[[7]]),
      q75 = as.numeric(parts[[8]]),
      q90 = as.numeric(parts[[9]]),
      q95 = as.numeric(parts[[10]]),
      variance = as.numeric(parts[[11]])
    )
  }

  parsed <- purrr::map(beta_lines, parse_line)
  parsed <- parsed[!vapply(parsed, is.null, logical(1))]
  if (!length(parsed)) tibble() else bind_rows(parsed)
}

transform_annotation <- function(q, boundary_eps = 1e-4, winsor_limit = 2.5) {
  q <- as.numeric(q)
  q_star <- pmin(pmax(q, -1 + boundary_eps), 1 - boundary_eps)
  a_raw <- stats::qnorm((q_star + 1) / 2)
  a_clip <- pmin(pmax(a_raw, -winsor_limit), winsor_limit)
  a_rms <- rms(a_clip)
  a <- if (!is.finite(a_rms) || a_rms == 0) rep(0, length(a_clip)) else a_clip / a_rms
  tibble(q_star = q_star, a_raw = a_raw, a_clip = a_clip, a = a, a_rms = a_rms)
}

PROJECT_ROOT <- find_project_root()
WORK_ROOT <- normalizePath(file.path(PROJECT_ROOT, "..", ".."), winslash = "/", mustWork = TRUE)
SUSINE_REPO <- file.path(WORK_ROOT, "SuSiNE - 2.0", "susine")
if (!dir.exists(SUSINE_REPO)) {
  stop("Could not locate susine repo at ", SUSINE_REPO)
}

devtools::load_all(SUSINE_REPO, quiet = TRUE)
library(susine)

MANIFEST_PATH <- file.path(PROJECT_ROOT, "config", "loci_manifest_sample_100_per_chrom.csv")
SELECTION_PATH <- resolve_selection_path(PROJECT_ROOT)
SELECTION_NAME <- tools::file_path_sans_ext(basename(SELECTION_PATH))
OUTPUT_DIR <- file.path(PROJECT_ROOT, "output")
LD_DIR <- file.path(OUTPUT_DIR, "ld")
ANNOTATION_DIR <- file.path(OUTPUT_DIR, "annotation", "alphagenome")
STEP5_DIR <- file.path(OUTPUT_DIR, "susine_mu0", SELECTION_NAME)
STEP6_DIR <- file.path(OUTPUT_DIR, "susine_baseline_validation", SELECTION_NAME)
dir.create(STEP6_DIR, recursive = TRUE, showWarnings = FALSE)

BASELINE_SUMMARY_PATH <- file.path(STEP5_DIR, "mu0_locus_summary.csv")
if (!file.exists(BASELINE_SUMMARY_PATH)) {
  stop("Step 5 baseline summary not found: ", BASELINE_SUMMARY_PATH)
}

SUSINE_L <- 10
SUSINE_SIGMA_0_2 <- 0.2
SUSINE_MAX_ITER <- 100
SUSINE_TOL <- 1e-5

cat("Project root:", PROJECT_ROOT, "\n")
cat("susine repo:", SUSINE_REPO, "\n")
cat("Step 5 dir:", STEP5_DIR, "\n")
cat("Step 6 dir:", STEP6_DIR, "\n")
```

## Scope

This workbook runs one baseline `susine::susine_rss()` model per selected locus using:

- Phase 1 `z`, `R`, and `n`
- the baseline per-locus annotation scale `baseline_c_l` from Step 5
- `mu_0 = baseline_c_l * a`

It captures the in-model beta-hat diagnostics printed during fitting and compares those diagnostics to the external beta-hat and baseline-annotation distributions built from the same summary inputs.

```{r load-selection}
manifest_df <- readr::read_csv(MANIFEST_PATH, show_col_types = FALSE)
selection_df <- readr::read_csv(SELECTION_PATH, show_col_types = FALSE)
baseline_summary_df <- readr::read_csv(BASELINE_SUMMARY_PATH, show_col_types = FALSE)

selected_loci_df <- merge_manifest_selection(manifest_df, selection_df) %>%
  distinct(locus_id, .keep_all = TRUE) %>%
  inner_join(
    baseline_summary_df %>% select(locus_id, baseline_c_l),
    by = "locus_id"
  )

if ("priority" %in% names(selected_loci_df)) {
  selected_loci_df <- selected_loci_df %>% arrange(priority, locus_id)
} else {
  selected_loci_df <- selected_loci_df %>% arrange(locus_id)
}

selected_loci_df %>%
  select(any_of(c("locus_id", "gene_name", "priority", "baseline_c_l")))
```

```{r run-baseline-validation}
run_rows <- list()
external_rows <- list()
internal_rows <- list()
compare_rows <- list()
error_rows <- list()

for (i in seq_len(nrow(selected_loci_df))) {
  locus_row <- selected_loci_df[i, ]
  locus_id <- locus_row$locus_id[[1]]
  gene_name <- locus_row$gene_name[[1]]
  baseline_c_l <- as.numeric(locus_row$baseline_c_l[[1]])

  locus_dir <- file.path(LD_DIR, locus_id)
  annotation_locus_dir <- file.path(ANNOTATION_DIR, locus_id)
  master_path <- file.path(locus_dir, sprintf("%s_phase1_master_variants.csv", gene_name))
  order_path <- file.path(locus_dir, sprintf("%s_LD_variant_order.tsv", gene_name))
  ld_long_path <- file.path(locus_dir, sprintf("%s_phase1_LD_R_long.parquet", gene_name))
  annotation_path <- file.path(annotation_locus_dir, sprintf("%s_alphagenome_filtered_scores.parquet", gene_name))

  cat(sprintf("[%d/%d] %s (%s)\n", i, nrow(selected_loci_df), locus_id, gene_name))

  missing_paths <- c(
    if (!file.exists(master_path)) master_path,
    if (!file.exists(order_path)) order_path,
    if (!file.exists(ld_long_path)) ld_long_path,
    if (!file.exists(annotation_path)) annotation_path
  )
  if (length(missing_paths) > 0) {
    error_rows[[length(error_rows) + 1L]] <- tibble(
      locus_id = locus_id,
      gene_name = gene_name,
      status = "missing_inputs",
      error_message = paste(missing_paths, collapse = " | ")
    )
    next
  }

  tryCatch({
    master_df <- readr::read_csv(master_path, show_col_types = FALSE) %>%
      filter(.data$ld_included) %>%
      transmute(
        variant_id,
        z_score,
        slope,
        slope_se,
        sample_size
      )

    order_df <- readr::read_tsv(order_path, show_col_types = FALSE) %>%
      arrange(.data$index)

    annotation_df <- arrow::read_parquet(annotation_path) %>%
      as_tibble() %>%
      group_by(.data$source_variant_id) %>%
      summarise(
        quantile_score = mean(.data$quantile_score, na.rm = TRUE),
        .groups = "drop"
      )

    aligned_df <- order_df %>%
      left_join(master_df, by = c("id" = "variant_id")) %>%
      left_join(annotation_df, by = c("id" = "source_variant_id")) %>%
      arrange(.data$index) %>%
      mutate(
        quantile_score = if_else(is.na(.data$quantile_score), 0, .data$quantile_score)
      ) %>%
      bind_cols(transform_annotation(.$quantile_score)) %>%
      mutate(
        adj = (.data$sample_size - 1) / (.data$z_score^2 + .data$sample_size - 2),
        beta_hat_std = sqrt(.data$adj) * .data$z_score / sqrt(.data$sample_size - 1),
        baseline_mu0_l = baseline_c_l * .data$a
      )

    if (any(is.na(aligned_df$z_score))) {
      stop("Aligned z_score vector contains NA values.")
    }
    if (!all(aligned_df$index == (seq_len(nrow(aligned_df)) - 1L))) {
      stop("LD order index is not contiguous from 0 to p-1.")
    }

    ld_long_df <- arrow::read_parquet(ld_long_path) %>% as_tibble()
    p <- nrow(aligned_df)
    R <- build_dense_ld(ld_long_df, p)

    if (max(abs(R - t(R))) > 1e-8) {
      stop("Reconstructed LD matrix is not symmetric within tolerance.")
    }
    if (max(abs(diag(R) - 1)) > 1e-8) {
      stop("Reconstructed LD matrix diagonal is not all ones within tolerance.")
    }

    n_rss <- round(mean(aligned_df$sample_size, na.rm = TRUE))

    external_beta_stats <- summarise_distribution(aligned_df$beta_hat_std) %>%
      rename_with(~ paste0("external_beta_hat_", .x))
    external_mu0_stats <- summarise_distribution(aligned_df$baseline_mu0_l) %>%
      rename_with(~ paste0("external_mu0_", .x))

    external_rows[[length(external_rows) + 1L]] <- bind_cols(
      tibble(
        locus_id = locus_id,
        gene_name = gene_name,
        baseline_c_l = baseline_c_l,
        n_variants = p,
        n_rss = n_rss
      ),
      external_beta_stats,
      external_mu0_stats
    )

    fit_log <- capture.output({
      fit <- susine::susine_rss(
        L = SUSINE_L,
        z = aligned_df$z_score,
        R = R,
        n = n_rss,
        mu_0 = baseline_c_l * aligned_df$a,
        sigma_0_2 = SUSINE_SIGMA_0_2,
        prior_update_method = "none",
        max_iter = SUSINE_MAX_ITER,
        tol = SUSINE_TOL,
        verbose = TRUE,
        verbose_beta_hat = TRUE,
        record_beta_hat_diagnostics = TRUE
      )
    })

    writeLines(fit_log, file.path(STEP6_DIR, sprintf("%s_%s_beta_hat_diagnostics.log", gene_name, locus_id)))

    diag_df <- extract_beta_hat_diagnostics(fit, fit_log) %>%
      mutate(
        locus_id = locus_id,
        gene_name = gene_name,
        baseline_c_l = baseline_c_l
      ) %>%
      relocate(.data$locus_id, .data$gene_name, .data$iteration, .data$effect)

    if (nrow(diag_df) == 0) {
      stop("No beta_hat diagnostics were returned or parsed from the fit log.")
    }

    internal_rows[[length(internal_rows) + 1L]] <- diag_df

    final_iter <- max(diag_df$iteration, na.rm = TRUE)
    final_diag_df <- diag_df %>%
      filter(.data$iteration == !!final_iter) %>%
      rename_with(~ paste0("internal_beta_hat_", .x), c(mean, median, mode, q25, q75, q90, q95, variance)) %>%
      mutate(final_iteration = final_iter)

    compare_rows[[length(compare_rows) + 1L]] <- final_diag_df %>%
      left_join(external_rows[[length(external_rows)]], by = c("locus_id", "gene_name", "baseline_c_l"))

    run_rows[[length(run_rows) + 1L]] <- tibble(
      locus_id = locus_id,
      gene_name = gene_name,
      baseline_c_l = baseline_c_l,
      n_variants = p,
      n_rss = n_rss,
      elbo_last = tail(fit$model_fit$elbo, 1),
      sigma_2_last = tail(fit$model_fit$sigma_2, 1),
      cs_count = length(fit$model_fit$PIPs[fit$model_fit$PIPs > 0]),
      max_pip = max(fit$model_fit$PIPs, na.rm = TRUE),
      final_iteration = final_iter,
      status = "completed",
      error_message = NA_character_
    )
  }, error = function(e) {
    error_rows[[length(error_rows) + 1L]] <<- tibble(
      locus_id = locus_id,
      gene_name = gene_name,
      baseline_c_l = baseline_c_l,
      status = "failed",
      error_message = conditionMessage(e)
    )
  })
}

baseline_run_summary_df <- bind_rows(run_rows)
external_expected_df <- bind_rows(external_rows)
internal_beta_hat_diagnostics_df <- bind_rows(internal_rows)
internal_vs_external_final_df <- bind_rows(compare_rows)
baseline_errors_df <- bind_rows(error_rows)

write_csv(baseline_run_summary_df, file.path(STEP6_DIR, "baseline_susine_run_summary.csv"))
write_csv(external_expected_df, file.path(STEP6_DIR, "baseline_external_expected_summary.csv"))
write_csv(internal_beta_hat_diagnostics_df, file.path(STEP6_DIR, "baseline_internal_beta_hat_diagnostics.csv"))
write_csv(internal_vs_external_final_df, file.path(STEP6_DIR, "baseline_internal_vs_external_final_iteration.csv"))
write_csv(baseline_errors_df, file.path(STEP6_DIR, "baseline_susine_errors.csv"))

baseline_run_summary_df
```

## External expected distributions

```{r external-summary}
external_expected_df
```

## Internal beta-hat diagnostics from the fitted model

```{r internal-summary}
internal_beta_hat_diagnostics_df %>%
  arrange(.data$locus_id, .data$iteration, .data$effect)
```

## Final-iteration comparison: external expectation vs in-model beta_hat

```{r compare-summary}
internal_vs_external_final_df %>%
  arrange(.data$locus_id, .data$effect)
```

## Failed loci

```{r failed-loci}
baseline_errors_df
```

## Comparison plots

```{r plots, fig.width=11, fig.height=7}
mean_plot <- internal_vs_external_final_df %>%
  ggplot(aes(x = .data$external_beta_hat_mean, y = .data$internal_beta_hat_mean, color = factor(.data$effect))) +
  geom_point(size = 2.2, alpha = 0.8) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "#444444") +
  facet_wrap(~ locus_id, scales = "free") +
  labs(
    title = "Final in-model beta_hat mean vs external expected beta_hat mean",
    x = "External beta_hat_std mean",
    y = "Internal beta_hat mean",
    color = "Effect"
  ) +
  theme_minimal(base_size = 12)

q95_plot <- internal_vs_external_final_df %>%
  ggplot(aes(x = .data$external_beta_hat_q95, y = .data$internal_beta_hat_q95, color = factor(.data$effect))) +
  geom_point(size = 2.2, alpha = 0.8) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "#444444") +
  facet_wrap(~ locus_id, scales = "free") +
  labs(
    title = "Final in-model beta_hat q95 vs external expected beta_hat q95",
    x = "External beta_hat_std q95",
    y = "Internal beta_hat q95",
    color = "Effect"
  ) +
  theme_minimal(base_size = 12)

mean_plot / q95_plot
```

## Notes

- This workbook uses the per-locus baseline `baseline_c_l` from Step 5 and does not pool scales across loci.
- The model is run with `verbose_beta_hat = TRUE` and `record_beta_hat_diagnostics = TRUE`.
- Console diagnostics are also written to one log file per locus under `output/susine_baseline_validation/<selection_name>/`.
